---
title: "Cluster identification - Giulia Beccaceci scRNA from stromal Il1r1 KO and epithelial Bmpr1a KO models; epithelial cells only. CCadj"
author: "Hilmar Berger"
output: 
 html_document:
    toc: true
    toc_depth: 3
    toc_float: false
    number_sections: true
    code_folding: hide
    
pdf_document:
    fig_caption: true
date: "`r format(Sys.time(), '%d %B, %Y %H:%M:%S')`"
---


```{r, message=FALSE}
rm(list=ls())
library(Seurat)
library(ggplot2)

library(SummarizedExperiment)
library(readxl)
library(writexl)
library(knitr)
library(data.table)
library(cowplot)
library(magrittr)
library(dplyr)

library(pheatmap)
library(homologene)

library(grid)
library(png)
library(speckle)
library(knitr)

result_folder = "../../Results/scRNA/antral_only"
if (!file.exists(result_folder)) dir.create(result_folder, recursive=T)

data_storage_folder = "../../Data/scRNA/Processed"
if (!file.exists(data_storage_folder)) dir.create(data_storage_folder, recursive=T)

library(future)
options("future.globals.maxSize"=2**33)
plan("sequential")

source("../Rutils/Dotplot_scRNA_v2.2.R")

rerun_dge_analyses = T
```

# General

We use preprocessed data that included only subpopulations determined to be epithelial, no doublets. We used variable features only from WT/untreated cells in order to not bias clusters by treatment effects.


```{r}
scData.combined <- readRDS(file.path(data_storage_folder, "Giulia_Il1r1_Bmpr1a_Harmony_Epi_only_v5_CCadj.rds"))
set.seed(20260709)
scData.combined <- FindClusters(scData.combined, resolution = 0.2)

DefaultAssay(scData.combined) <- "RNA"
scData.combined <- JoinLayers(scData.combined)

set.seed(1234)
```

# Cluster overview

```{r, fig.width=8, fig.height=5}
DimPlot(scData.combined, reduction = "umap", label = TRUE)
```

```{r, fig.width=10, fig.height=5}
DimPlot(scData.combined, reduction = "umap", label = TRUE, split.by = "group", ncol = 4)
```


# Final cluster assigment

```{r}
cluster_assignment = read.table("Cluster_assignments/Cluster_assignment_Epi_only_2026-07-15_v5.1.txt", sep="\t",header=T, stringsAsFactors = F)
options(width = 180)
print(cluster_assignment)
rownames(cluster_assignment) = as.character(cluster_assignment$ClusterID)
scData.combined = AddMetaData(scData.combined, cluster_assignment[as.character(scData.combined$seurat_clusters),"Label"], col.name = "cluster_label")
scData.combined = AddMetaData(scData.combined, cluster_assignment[as.character(scData.combined$seurat_clusters),"Type"], col.name = "cluster_type")

scData.combined$cluster_label <- factor(scData.combined$cluster_label, levels = sort(unique(cluster_assignment$Label)))
```


```{r}
group_labels <- c("epiBmpr1aKO uninfected" = "Bmpr1a KO","epiBmpr1aWT uninfected"="Bmpr1a WT","Stromal IL1r1 WT Hpy 2m inf"="H.pylori 2m","Stromal IL1r1 WT uninfected"= "Uninfected")

scData.combined@meta.data$group_label <- group_labels[as.character(scData.combined$group)]
scData.combined@meta.data$Experiment <- ifelse(scData.combined$group_label %in% c("Uninfected","H.pylori 2m"), "Hpy infection", "Bmpr1a KO") |> factor(levels=c("Hpy infection","Bmpr1a KO"))
```


```{r}
group_order = c("Uninfected","H.pylori 2m", "Bmpr1a WT",  "Bmpr1a KO" )
cluster_order <- c("Stem","TA/Progenitor","Pit","Surface mucous","Intestinal-like","stressed cells")
```

```{r}
scData.combined$cluster_label <- factor(scData.combined$cluster_label, levels = cluster_order)
scData.combined$group_label <- factor(scData.combined$group_label, levels = group_order)
```

```{r}
scData.combined = SetIdent(scData.combined, value="cluster_label" )
```



```{r, fig.width=10, fig.height=5}
p = DimPlot(scData.combined, reduction = "umap", label = TRUE) + NoLegend()
print(p)
```

```{r, fig.width=15, fig.height=5}
p = DimPlot(scData.combined, reduction = "umap", label = T, split.by="group") + NoLegend()
print(p)
```


## Distribution of clusters within samples

```{r}
m = table(scData.combined$cluster_label, scData.combined$sampleID)
kable(m)
```

```{r, fig.width=10,fig.height=7}
mm  = as.matrix.data.frame(m)
colnames(mm) = colnames(m)
rownames(mm) = rownames(m)

mm_r = sweep(mm, 2, apply(mm,2,sum), "/")

pheatmap(mm_r, cluster_cols = F, cluster_rows=T, main="Proportion of each cluster per sample")
```


## Distribution of clusters within conditions

```{r}
m = with(subset(scData.combined@meta.data, !cluster_label %in% c("stressed cells")), table(droplevels(cluster_label), group_label))
m
```

```{r}
mm  = as.matrix.data.frame(m)
colnames(mm) = colnames(m)
rownames(mm) = rownames(m)

mm_r = sweep(mm, 2, apply(mm,2,sum), "/")

options(width=200)
kable(round(mm_r,3))
```


```{r, fig.width=6,fig.height=5}
pheatmap(mm_r, cluster_cols = F, cluster_rows=T)
```

```{r, fig.width=14, fig.height=8}
mm_rel_ts = reshape2::melt(t(mm_r))
colnames(mm_rel_ts) = c("Group","Cluster","Proportion_cells")
mm_rel_ts$Cluster = factor(paste0("",mm_rel_ts$Cluster))

ggplot(mm_rel_ts, aes(x=Group, y = Proportion_cells, group=Cluster)) + geom_bar(stat="identity") + theme(axis.text.x = element_text(angle=60, hjust=1)) + facet_wrap(~ Cluster )
```

```{r, fig.width=10, fig.height=6}
#cluster_order = sort(unique(mm_rel_ts$Cluster))
#group_order = c("epiBmpr1aWT uninfected","epiBmpr1aKO uninfected","Stromal IL1r1 WT uninfected","Stromal IL1r1 WT Hpy 2m inf")
mm_rel_ts$Cluster = factor(mm_rel_ts$Cluster, levels=cluster_order)
mm_rel_ts$Group <- factor(mm_rel_ts$Group,group_order)
ggplot(mm_rel_ts, aes(x=Group, y = Proportion_cells, fill=Cluster)) + geom_bar(stat="identity") + 
  theme(axis.text.x = element_text(angle=45, hjust=1))
```

```{r, fig.width=12, fig.height=5}
m = with(subset(scData.combined@meta.data, !cluster_label %in% c("stressed cells")),table(droplevels(cluster_label), orig.ident) )
aa <- prop.table(m, margin=2) |> reshape2::melt()
aa <- merge(aa, unique(scData.combined@meta.data[, c("orig.ident","group_label")]), by.x="orig.ident", by.y="orig.ident")
aa$group <- factor(aa$group_label, levels=group_order)
aa$cluster = factor(aa$Var1, levels=cluster_order)
ggplot(aa, aes(x=cluster, fill=group, y=value)) + geom_boxplot() + ggtitle("Clusters in groups") + ylab("Average proportion in samples") +  theme(axis.text.x = element_text(angle=45, hjust = 1, size=14))
```

## Comparison of celltype composition

We here use the logit-transformation from R package **speckle** to compare cluster proportions between groups.

```{r, results='asis'}
meta <- subset(scData.combined@meta.data, !cluster_label %in% c("stressed cells"))
m_axin2 <- subset(meta, genotype %in% c("Axin2CreErt2-Bmpr1afl/fl KO","Axin2CreErt2-Bmpr1afl/fl WT"))
res_axin2 <- speckle::propeller(cluster = droplevels(m_axin2$cluster_label), 
                          sample = m_axin2$orig.ident, 
                          group = as.character(m_axin2$group_label))

m_col1 <- subset(meta, genotype %in% c("Col1a2CreErt2-Il1r1fl/fl WT"))
res_col1 <- speckle::propeller(cluster = droplevels(m_col1$cluster_label), 
                          sample = m_col1$orig.ident, 
                          group = as.character(m_col1$group_label))

kable(res_axin2)
kable(res_col1)
```



```{r, fig.width=8, fig.height=5}
m = with(subset(scData.combined@meta.data, !cluster_label %in% c("stressed cells")),table(cluster_label |> droplevels(), orig.ident))
aa <- prop.table(m, margin=2) |> reshape2::melt()
aa <- merge(aa, unique(scData.combined@meta.data[, c("orig.ident","group_label","Experiment")]), by.x="orig.ident", by.y="orig.ident")
aa$group <- factor(aa$group, levels=group_order)

aa$xx <- ifelse(aa$group %in% c("Uninfected","Bmpr1a WT"), 1,2)
aa$cluster = factor(gsub(" |/" , "\n", aa$Var1), levels=gsub(" |/", "\n", cluster_order))
p <- ggplot(aa, aes(x=xx, col=group, y=value)) + geom_point() + stat_summary(geom = "crossbar", fun = mean, width=0.5) + ggtitle("Epithelial subcluster proportions across groups - this study") + ylab("Average proportion in samples") + 
  theme(axis.text.x = element_blank(), axis.ticks.x=element_blank(), strip.text = element_text(size=12)) + facet_grid(Experiment ~ cluster, scales = "free_x") + xlab("")

p
```

```{r}
saveRDS(scData.combined@meta.data,file=file.path(data_storage_folder, "Giulia_Il1r1_Bmpr1a_Harmony_epi_Metadata_v5.1_CCadj.rds"))
```

# Marker genes for clusters (assessed on WT cells only)

```{r}
if(rerun_dge_analyses) {
  cluster_markers <- FindAllMarkers(object = scData.combined, only.pos = TRUE, assay = "RNA")
  cluster_markers$diff_pct <- cluster_markers$pct.1 - cluster_markers$pct.2
  saveRDS(cluster_markers, file=file.path(data_storage_folder, "All_cluster_markers_Epi_only_v5.1_CCadj.rds"))
  write_xlsx(cluster_markers,file.path(result_folder, "All_cluster_markers_Epi_only_v5.1_CCadj.xlsx"))
} else {
  cluster_markers <- readRDS(file.path(data_storage_folder, "All_cluster_markers_Epi_only_v5.1_CCadj.rds"))
}
```

```{r}
plot_top_genes <- function(data, dge_results_all, title) {
  for (n in sort(unique(dge_results_all$cluster))) {
    tmp = subset(dge_results_all, cluster==n)
    if(nrow(tmp)<1) next
    tmp_up = subset(tmp, avg_log2FC > 0)
    if(nrow(tmp_up)<1) next
    tmp_up$direction = "up"
    #tmp_down = subset(tmp, avg_log2FC < 0)
    #tmp_down$direction = "down"
    
    #final_top_genes = rbind(head(tmp_up,10), head(tmp_down,10))
    final_top_genes = head(tmp_up, 10)
    final_top_genes = final_top_genes[order(final_top_genes$avg_log2FC),]
    final_top_genes$gene = factor(final_top_genes$gene, levels = final_top_genes$gene)
    p = ggplot(final_top_genes, aes(y=avg_log2FC, x=gene, fill=direction)) + geom_bar(stat="identity") + ggtitle(paste0(n, ", ", title)) + coord_flip() + theme(axis.text.y =  element_text(size=12) )
    print(p)
    
  }
}
```


## Percent expression of top markers across clusters

```{r, fig.width=16, fig.height=14}
top_5_markers = unique(unlist(tapply(cluster_markers$gene, cluster_markers$cluster, head, 5)))
DotPlot(scData.combined, features = top_5_markers , dot.scale = 4, cols = c("red", "blue"), assay = "RNA") +    RotatedAxis()
```



## Log2 Fold Changes of top 10 DE genes from each cluster

```{r, fig.width=7, fig.height=4}
plot_top_genes(scData.combined, cluster_markers, "Cluster markers")
```




```{r}

plot_genes <- function(dataset, genes) {
  
  all_genes =rownames(dataset@assays$RNA)
  
  for (g in genes) {
    if (!g %in% all_genes) {
      print(paste0("Gene ",g," has no data (possibly filtered out)."))
      next
    } else {
        p1  =FeaturePlot(object = dataset, features = g, reduction="umap", label = T)
        p2 = VlnPlot(dataset, features = g, ncol = 1, pt.size = 0.01)
        print(plot_grid(p1,p2, ncol=2,align="h"))
    }
  }
}
```


# Individual marker expression

## Global markers 

These are global cell markers for major cell lineages and types (B/T-cells, Mo/MF etc)

CD45 is Ptprc.  
Pglyrp1 - circulating polymorphonuclear leukocytes (granulocytes + mast cells)  

We here add Cd3e/d/g as pan-T marker and Cd20 (Ms4a1) as mature B-cell marker.
Gata3, Cxcr6, Rora - T-cell markers

Ly6a(Sca-1) - positive in HSC/MPP
Procr,Fdg5, Hoxb5 - HSC

Epor, Slamf1(Cd150) - MegE - lineage

```{r, fig.width=14, fig.height=6}

sel_markers=c("Ptprc","Pglyrp1","Cd4","Cd3e","Cd3d","Cd3g", "Gata3","Cxcr6","Rora","Cd7","Cd8a","Cd19","Ms4a1","Jchain","Aif1","Cd68","Flt3","Ly6a", "Cd34","Procr","Fgd5","Hoxb5","Epor","Slamf1","Vim","Pecam1","Lyve1","Vwf","Col1a1","Myh11","Acta2","Epcam","Krt20","Lgr5","Axin2","S100b","Ncam1","Sox2","Gfap")
DefaultAssay(scData.combined) <- "RNA"

#plot_genes(scData.combined, sel_markers)
DotPlot(scData.combined, features=sel_markers, cols="RdYlBu", assay="RNA")+ theme(axis.text.x=element_text(angle=45, hjust=1))

```

## Granulocyte subpopulations

Pglyrp1 - circulating polymorphonuclear leukocytes (granulocytes + mast cells)  
Csf3r - colony stimulating factor 3 receptor (granulocytes)  G-CSF-R. Higher in mature blood neutrophils (Shinjo et al. 1995 doi: 10.1111/j.1365-2141.1995.tb05390.x)

Mpo,Elane - neutrophil markers  
Prg2 - eosinophils (Dahlin)  
Gzmb, Cma1 - Mast cell (Dahlin)  
Ms4a2, Cpa3 Mast cell, basophil (Dahlin)  
Fcer1a - basophils, mastcells - high affinity IgE receptor  
Ggnt2 - eosinophils  
Il5ra - eosinophil progenitor  



```{r, fig.width=14, fig.height=6}

sel_markers=c("Pglyrp1","Csf3r","Mpo","Elane","Gngt2","Il5ra","Prg2","Gzmb","Cma1","Ms4a2","Cpa3","Fcer1a","Cxcr1","Cxcr2")
DefaultAssay(scData.combined) <- "RNA"

#plot_genes(scData.combined, sel_markers)
DotPlot(scData.combined, features=sel_markers, cols="RdYlBu", assay="RNA")+ theme(axis.text.x=element_text(angle=45, hjust=1))

```


## Monocyte and macrophage markers

Adgre1 = F4/80 = Emr1, Blood Monocytes  

CD14lowCD16high - non-classical MoMF    
CD16 - Fcgr4 (FCGR3A/B)  
Fcgr1 (Cd64) MoMF  
Spn (Cd45) classical Mo  
Cd11b = Itgam  

EMP (Mass et al.) Csf1r + Kit + CD45 low AA4.1(Cd93)+   

```{r, fig.width=14, fig.height=6}
sel_markers=c("Adgre1","Ccr2","Ccr5","Arg1","Mgst1","Fn1","Cd74","Cd68","S100a8","S100a9","Cd14","Fcgr3","Cx3cr1","Fcgr1a","Spn","Itgam","Csf1r","Kit","Cd93","Il7r","Flt3","Cxcr4","Il1b","Aif1","Timd4")
DefaultAssay(scData.combined) <- "RNA"

#plot_genes(scData.combined, sel_markers)
DotPlot(scData.combined, features=sel_markers, cols="RdYlBu", assay="RNA")+ theme(axis.text.x=element_text(angle=45, hjust=1))
```


## DC markers

Xcr1 cDC  
Siglech pDC  
Irf8 pre-DC1  
Ccr7 cDC2  
Pclaf (2810417H13Rik) CMoP / pre-DC1  
Cd209a cDC2, pDC  
Tlr2 cDC  
Tlr4 cDC  

Socs2 Tissue migratory DC (Nirschl et al. 2017, 10.1016/j.cell.2017.06.016)  


```{r, fig.width=14, fig.height=6}
sel_markers=c( "Socs2", "Xcr1", "Siglech", "Cd209a", "Irf8", "Pclaf", "Tlr2","Tlr4","Tlr7","Tlr9","CD1C","CLEC4C")
DefaultAssay(scData.combined) <- "RNA"

#plot_genes(scData.combined, sel_markers)
DotPlot(scData.combined, features=sel_markers, cols="RdYlBu", assay="RNA")+ theme(axis.text.x=element_text(angle=45, hjust=1))
```


## MHC-I/II

```{r, fig.width=14, fig.height=6}
sel_markers=c("H2-Oa","H2-DMa","H2-DMb1","H2-DMb2","H2-Ob","H2-Ab1","H2-Aa","H2-Eb1","H2-Eb2")
#sel_markers = c("HLA-A","HLA-B","HLA-C","HLA-D","HLA-E","HLA-F","HLA-G","HLA-DRA","HLA-DRB","HLA-DQA1","HLA-DQB1","HLA-DPA1","HLA_DPB1","HLA_DMA","HLA_DMB","HLA-DOA","HLA_DOB","MICA","MICB")
DefaultAssay(scData.combined) <- "RNA"

#plot_genes(scData.combined, sel_markers)
DotPlot(scData.combined, features=sel_markers, cols="RdYlBu", assay="RNA")+ theme(axis.text.x=element_text(angle=45, hjust=1))
```

## Stomach site markers

```{r, fig.height=8, fig.width=12}
dotplot_scRNA_multi_feature(scData.combined, features=c("Apc","Axin2","Anxa10","Sox2","Gata4","Meis1","Pdx1","Gata6"),  group="group", cluster="cluster_label", scale=T)+ theme(axis.text.x=element_text(angle=45, hjust=1))
```

```{r}
sd1 <- subset(scData.combined, group %in% c("epiBmpr1aWT uninfected","Stromal IL1r1 WT uninfected"))
```


```{r, fig.height=8, fig.width=10}
dotplot_scRNA_multi_feature(sd1, features=c("Apc","Axin2","Anxa10","Sox2","Gata4","Meis1","Pdx1","Gata6"),  group="tissue_site", cluster="cluster_label", scale=T)+ theme(axis.text.x=element_text(angle=45, hjust=1))
```



```{r, fig.width=10, fig.height=5}
for (gg in c("Apc","Anxa10","Sox2","Gata6","Gata4","Meis1","Pdx1","Dmbt1","Cblif")) {
  p <- FeaturePlot(sd1, features = gg, order=T, split.by="tissue_site") + NoLegend()
  print(p)
}
```

```{r}
aa <- LayerData(scData.combined, layer = "data", features=c("Pdx1","Sox2"))
plot(density(aa["Pdx1",])); abline(v=0.2, col="red")
barplot(tapply(aa["Pdx1",], scData.combined$seurat_clusters, mean), las=2); abline(h=0.2, col="red")

plot(density(aa["Sox2",])); abline(v=0.15, col="red")
barplot(tapply(aa["Sox2",], scData.combined$seurat_clusters, mean), las=2); abline(h=0.15, col="red")
```

## Stomach epithelial cell markers

```{r, fig.width=14, fig.height=4}
sel_markers=c("Lgr5","Ascl2","Aqp5","Muc5ac","Muc6","Krt20","Gkn1","Pgc","Pga5", "Cblif","Mki67","Mcm3","Dclk1",  "Atp4a","Atp4b","Sox2","Cdx2","Muc2","Olfm4", "Aldh1a1","Tff1","Tff2","Chga","Chgb","Tph1","Sst","Maoa","Maob","Gast","Neurog3","Hdc","Grpr","Dach1","Cntfr","Slc5a5","Adh1c") # |> human2mouse(db=homologeneData2) |> (\(x) x$mouseGene[!is.na(x$mouseGene)] |> unique())()
DefaultAssay(scData.combined) <- "RNA"

#plot_genes(scData.combined, sel_markers)
DotPlot(scData.combined, features=sel_markers, cols="RdYlBu", assay="RNA")+ theme(axis.text.x=element_text(angle=45, hjust=1))
```

```{r, fig.width=14, fig.height=4}

sel_markers=c("Tff1","Tff2","Muc6","Muc5ac","Gkn1","Cblif","Lgr5","Axin2","Aqp5","Cd44","mt-Nd1","Muc4","Mki67","Mcm3",
              "Sptssb","Aldh3a1","Cd74","Duox2", "H2-Ab1",
              "Krt20",
              "Fos","Jun","Atf3","Cdx2","Muc2","Olfm4")  

#plot_genes(scData.combined, sel_markers)
p <- DotPlot(scData.combined, features=sel_markers, cols="RdYlBu", assay="RNA", group.by="cluster_label")+ theme(axis.text.x=element_text(angle=45, hjust=1)) + labs(x="", y="" , title="Selected markers for epithelial cell populations - this study")

p
```

```{r, fig.width=14, fig.height=4}
# taken from https://doi.org/10.1016/j.cels.2016.09.002
# Slc18a2 = Vmat2
DotPlot(scData.combined, c("Pdx1","Sox2","Ascl2","Aqp5","Lgr5","Axin2","Lrig1","Bmi1","Bhlha15","Stmn1","Iqgap3","Tff1","Tff2","Tnfrsf19","Cblif","Chia1","Pga5","Pgc","Atp4a","Muc6","Muc5ac","Krt20","Mki67","Cd74","Ifi47","Ly6a"), cols="RdYlBu",assay="RNA", col.max = 1.5, col.min=-1.5) + theme(axis.text.x = element_text(angle=45, hjust=1))
```

## Other epithelial markers

```{r, fig.width=10, fig.height=4}
# ,"Klf4","Myc","Nanog"
DotPlot(scData.combined, c("Sox2","Hnf4a","Hnf4g","Atoh1","Cdx2","Krt5","Krt14","Krt7","Krt19","Krt20","Krt8","Krt18","Aqp8","Car4","Aqp4","Muc5ac","Muc6","Tff1","Tff2","Tff3","Ly6a","Vtn","Gc","Spink4","Muc2"), cols="RdYlBu",assay="RNA") + theme(axis.text.x = element_text(angle=45, hjust=1))
```


```{r, fig.width=10, fig.height=4}
# taken from https://doi.org/10.1016/j.cels.2016.09.002
# Slc18a2 = Vmat2
DotPlot(scData.combined, c("Gcg","Irx2","Arx","Fev","Gc","Ttr","Pgr","Mafb","Ins1","Mafa","Hhex","Sst","Isl1","Rbp4","Gip","Pyy","Etv1","Pax6","Gast","Pdx1","Cck","Nts","Hdc"), cols="RdYlBu",assay="RNA") + theme(axis.text.x = element_text(angle=45, hjust=1))
```

## WAE markers

SSC2c cluster markers from Ayyaz, A., Kumar, S., Sangiorgi, B. et al. Single-cell transcriptomes of the regenerating intestine reveal a revival stem cell. Nature 569, 121–125 (2019). https://doi.org/10.1038/s41586-019-1154-y  

and markers from WAE cells in Nakate et al. (DOI: 10.1126/scitranslmed.adg5252)

```{r, fig.width=8, fig.height=4}
sel_markers = c("Clu","Cd44","Ly6a","Aldh1a3","Cldn4","Krt14","Krt7","Anxa1","Anxa3")
DotPlot(scData.combined, features=sel_markers, cols="RdYlBu", assay="RNA")+ theme(axis.text.x=element_text(angle=45, hjust=1))
```

```{r, fig.width=22, fig.height=5}
sel_markers <- c("Lgr5","Aqp5","Muc5ac","Muc6","Gkn1","Pgc","Cblif","Mki67","Krt20","Olfm4", "Atp4a","Cdx2","Muc2","Tff2","Cd74","H2-Eb1","Bmpr1a","Il1r1","Ly6a","Anxa1","Anxa3")
dotplot_scRNA_multi_feature(scData.combined, cluster = "cluster_label", group = "group_label", features = sel_markers, scale=T, group_order = group_order) # 
```

<!-- ## Yap/fetal/RSC markers -->


<!-- ```{r} -->
<!-- sel_clusters <- unique(scData.combined$cluster_label) |> as.character() -->
<!-- sel_clusters <- sel_clusters[!sel_clusters %in% c("stressed cells")] -->

<!-- sd_bmp <- subset(scData.combined, genotype %in% c("Axin2CreErt2-Bmpr1afl/fl KO","Axin2CreErt2-Bmpr1afl/fl WT") & cluster_label %in% sel_clusters) -->
<!-- sd_bmp@meta.data$condition <- gsub("epiBmpr1a","", sd_bmp$group) |> gsub(" uninfected", "", x=_) -->

<!-- sd_inf <- subset(scData.combined, genotype %in% c("Col1a2CreErt2-Il1r1fl/fl WT") & cluster_label %in% sel_clusters) -->
<!-- sd_inf@meta.data$condition <- gsub("Stromal IL1r1 WT ","", sd_inf$group) -->
<!-- ``` -->


<!-- ```{r, fig.width=14, fig.height=4.5} -->
<!-- sel_markers <- c("Anxa1", "Ccn2", "Ccn1", "Wfdc2", "Ly6a", "Tubb6", "Sftpd", "Tacstd2", "Gja1","Lgr5", "Lgr4", "Bmi1", "Tff2",  "Hopx", "Smoc2", "Lrig1", "Vil1") -->

<!-- dotplot_scRNA_multi_feature(sd_bmp, cluster = "cluster_label", group = "condition", features = sel_markers, scale=T, group_order = c("WT","KO"), cluster_order = cluster_order) -->

<!-- dotplot_scRNA_multi_feature(sd_inf, cluster = "cluster_label", group = "condition", features = sel_markers, scale=T, group_order = c("uninfected","Hpy 2m inf"), cluster_order = cluster_order) -->

<!-- sel_markers_inf <- read.table(text="IL1A -->
<!-- TNF -->
<!-- IL1R1 -->
<!-- TNFRSF1A -->
<!-- TNFRSF1B -->
<!-- RELA -->
<!-- IKBKG -->
<!-- IKBKB -->
<!-- NFKB1 -->
<!-- MYD88 -->
<!-- TNFAIP3 -->
<!-- CHUK -->
<!-- MAP3K7 -->
<!-- MAP3K1 -->
<!-- MAP3K14 -->
<!-- TRADD -->
<!-- RIPK1 -->
<!-- FADD -->
<!-- TRAF6 -->
<!-- TAB1 -->
<!-- NFKBIA -->
<!-- ")$V1 |> unique() |> homologene::human2mouse(db = homologene::homologeneData2) |> (\(x) x$mouseGene)() |> unique() -->

<!-- dotplot_scRNA_multi_feature(sd_bmp, cluster = "cluster_label", group = "condition", features = sel_markers_inf, scale=T, group_order = c("WT","KO")) + ggtitle("BIOCARTA_NFKB_PATHWAY") # , cluster_order = cluster_order -->
<!-- ``` -->


<!-- ```{r, fig.width=14, fig.height=4} -->
<!-- sel_markers <- c("Tead4", "Wfdc2", "Itga2", "Tgm2", "Lrg1", "Anxa1", "Sftpd", "Ccn1", "Vnn1", "Clu", "Socs3", "Ly6a", "Il33", "Ly6d", "Gpnmb", "Ereg") -->

<!-- p1 <- dotplot_scRNA_multi_feature(sd_inf, cluster = "cluster_label", group = "condition", features = sel_markers, scale=T, group_order = c("uninfected","Hpy 2m inf"), cluster_order = cluster_order) + ggtitle("Regenerative state genes - +/- Hpy infection") + xlab("") + ylab("") -->
<!-- p1 -->
<!-- p2 <- dotplot_scRNA_multi_feature(sd_bmp, cluster = "cluster_label", group = "condition", features = sel_markers, scale=T, group_order = c("WT","KO"), cluster_order = cluster_order) + ggtitle("Regenerative state genes - KO vs. WT") + xlab("") + ylab("") -->
<!-- p2 -->
<!-- #  -->
<!-- # pdf(file.path(result_folder, "Selected_Regenerative_state_genes_Hpy_inf.pdf"), width=14, height = 6) -->
<!-- # print(p1) -->
<!-- # dev.off() -->
<!-- #  -->
<!-- # pdf(file.path(result_folder, "Selected_Regenerative_state_genes_Bmpr1a_KO.pdf"), width=14, height = 6) -->
<!-- # print(p2) -->
<!-- # dev.off() -->

<!-- ``` -->


<!-- ```{r, fig.width=8.5, fig.height=5} -->
<!-- sel_mat <- AverageExpression(sd_inf, assays = "RNA", features = sel_markers, group.by = c("condition","cluster_label"))[["RNA"]] -->

<!-- # sel_mat_ts <- reshape2::melt(as.matrix(sel_mat)) -->
<!-- # colnames(sel_mat_ts) <- c("Gene","group","avg_expression") -->
<!-- # tmp <- strsplit(as.character(sel_mat_ts$group), "_") -->
<!-- # sel_mat_ts$condition <- sapply(tmp, `[`, 1) |> factor(levels=c("uninfected","Hpy 2m inf")) -->
<!-- # sel_mat_ts$cluster <- sapply(tmp, `[`, 2) |> factor(levels=cluster_order) -->
<!-- # sel_mat_ts <- sel_mat_ts[order(sel_mat_ts$Gene, sel_mat_ts$cluster, sel_mat_ts$condition),] -->
<!-- # sel_mat_ts$group <- factor(sel_mat_ts$group, levels=unique(sel_mat_ts$group)) -->
<!-- #  -->
<!-- # tmp <- reshape2::dcast(sel_mat_ts, Gene ~ group, value.var="avg_expression") -->

<!-- # agg_mat <- as.matrix(tmp[, 2:ncol(tmp)]) -->
<!-- # rownames(agg_mat) <- tmp$Gene -->

<!-- agg_mat <- as.matrix(sel_mat) -->

<!-- tmp <- strsplit(colnames(agg_mat), "_") -->
<!-- col_anno <- data.frame(row.names=colnames(agg_mat),  -->
<!--                        Condition = sapply(tmp, `[`,1) |> factor(levels=c("uninfected","Hpy 2m inf")) ,  -->
<!--                        Cluster = sapply(tmp, `[`,2)|> factor(levels=cluster_order)) -->

<!-- agg_mat <- agg_mat[sel_markers, order(col_anno[match(colnames(agg_mat), rownames(col_anno)),"Cluster"],  -->
<!--                            col_anno[match(colnames(agg_mat), rownames(col_anno)),"Condition"])] -->

<!-- anno_cols <- list(Condition=c("uninfected"="lightgreen","Hpy 2m inf"="red")) -->

<!-- pheatmap(agg_mat, cluster_cols=F, cluster_rows = F, scale="row", gaps_col = seq(2, ncol(agg_mat)-2, 2) , breaks=c(-5,-3,seq(-2,2,4/96),3,5),  -->
<!--          annotation_col = col_anno[,c("Condition"), drop=F], -->
<!--          labels_col = col_anno[match(colnames(agg_mat), rownames(col_anno)), "Cluster"], -->
<!--          annotation_colors = anno_cols, -->
<!--          main="Scaled average expression of regenerative markers, Hpy vs. uninfected") -->
<!-- #  -->
<!-- # pheatmap(agg_mat, cluster_cols=F, cluster_rows = F, scale="row", gaps_col = seq(2, ncol(agg_mat)-2, 2) , breaks=c(-5,-3,seq(-2,2,4/96),3,5),  -->
<!-- #          annotation_col = col_anno[,c("Condition"), drop=F], -->
<!-- #          labels_col = col_anno[match(colnames(agg_mat), rownames(col_anno)), "Cluster"], -->
<!-- #          annotation_colors = anno_cols, -->
<!-- #          main="Scaled average expression of regenerative markers, Hpy vs. uninfected",  -->
<!-- #          filename = file.path(result_folder, "Regenerative_state_markers_Hpy_inf_Heatmap.pdf"), width = 8.5, height = 5) -->

<!-- ``` -->


<!-- ```{r, fig.width=8.5, fig.height=5} -->
<!-- sel_mat <- AverageExpression(sd_bmp, assays = "RNA", features = sel_markers, group.by = c("condition","cluster_label"))[["RNA"]] -->

<!-- agg_mat <- as.matrix(sel_mat) -->

<!-- tmp <- strsplit(colnames(agg_mat), "_") -->
<!-- col_anno <- data.frame(row.names=colnames(agg_mat),  -->
<!--                        Condition = sapply(tmp, `[`,1) |> factor(levels=c("WT","KO")) ,  -->
<!--                        Cluster = sapply(tmp, `[`,2)|> factor(levels=cluster_order)) -->

<!-- agg_mat <- agg_mat[sel_markers, order(col_anno[match(colnames(agg_mat), rownames(col_anno)),"Cluster"],  -->
<!--                            col_anno[match(colnames(agg_mat), rownames(col_anno)),"Condition"])] -->

<!-- anno_cols <- list(Condition=c("WT"="lightgreen","KO"="red")) -->

<!-- pheatmap(agg_mat, cluster_cols=F, cluster_rows = F, scale="row", gaps_col = seq(2, ncol(agg_mat)-2, 2) , breaks=c(-5,-3,seq(-2,2,4/96),3,5),  -->
<!--          annotation_col = col_anno[,c("Condition"), drop=F], -->
<!--          labels_col = col_anno[match(colnames(agg_mat), rownames(col_anno)), "Cluster"], -->
<!--          annotation_colors = anno_cols, -->
<!--          main="Scaled average expression of regenerative markers, Hpy vs. uninfected") -->

<!-- # pheatmap(agg_mat, cluster_cols=F, cluster_rows = F, scale="row", gaps_col = seq(2, ncol(agg_mat)-2, 2) , breaks=c(-5,-3,seq(-2,2,4/96),3,5),  -->
<!-- #          annotation_col = col_anno[,c("Condition"), drop=F], -->
<!-- #          labels_col = col_anno[match(colnames(agg_mat), rownames(col_anno)), "Cluster"], -->
<!-- #          annotation_colors = anno_cols, -->
<!-- #          main="Scaled average expression of regenerative markers, Bmpr1a KO vs. WT",  -->
<!-- #          filename = file.path(result_folder, "Regenerative_state_markers_Bmpr1a_KO_Heatmap.pdf"), width = 8.5, height = 5) -->
<!-- ``` -->



<!-- ```{r, fig.width=8, fig.height=4} -->
<!-- sel_markers_inf <- c("Cxcl1","Cxcl2","Cxcl5","Ccl20","Cxcl10","Ifit3","Isg20","Ifi47","Tnf") -->
<!-- p1 <- dotplot_scRNA_multi_feature(sd_inf, cluster = "cluster_label", group = "condition", features = sel_markers_inf, scale=T, group_order = c("uninfected","Hpy 2m inf"), cluster_order = cluster_order) + ggtitle("Inflammatory cytokines - +/- Hpy infection") + xlab("") + ylab("") -->
<!-- p1 -->

<!-- p2 <- dotplot_scRNA_multi_feature(sd_bmp, cluster = "cluster_label", group = "condition", features = sel_markers_inf, scale=T, group_order = c("WT","KO"), cluster_order = cluster_order) + ggtitle("Inflammatory cytokines - Bmpr1a KO vs. WT") + xlab("") + ylab("") -->
<!-- p2 -->
<!-- #  -->
<!-- # pdf(file.path(result_folder, "Inflammatory cytokines_Hpy_inf.pdf"), width=8, height = 6) -->
<!-- # print(p1) -->
<!-- # dev.off() -->
<!-- #  -->
<!-- # pdf(file.path(result_folder, "Inflammatory cytokines_Bmpr1a_KO.pdf"), width=8, height = 6) -->
<!-- # print(p2) -->
<!-- # dev.off() -->

<!-- ``` -->

<!-- ```{r, fig.width=8, fig.height=3} -->
<!-- sel_mat <- AverageExpression(sd_inf, assays = "RNA", features = sel_markers_inf, group.by = c("condition","cluster_label"))[["RNA"]] -->

<!-- # sel_mat_ts <- reshape2::melt(as.matrix(sel_mat)) -->
<!-- # colnames(sel_mat_ts) <- c("Gene","group","avg_expression") -->
<!-- # tmp <- strsplit(as.character(sel_mat_ts$group), "_") -->
<!-- # sel_mat_ts$condition <- sapply(tmp, `[`, 1) |> factor(levels=c("uninfected","Hpy 2m inf")) -->
<!-- # sel_mat_ts$cluster <- sapply(tmp, `[`, 2) |> factor(levels=cluster_order) -->
<!-- # sel_mat_ts <- sel_mat_ts[order(sel_mat_ts$Gene, sel_mat_ts$cluster, sel_mat_ts$condition),] -->
<!-- # sel_mat_ts$group <- factor(sel_mat_ts$group, levels=unique(sel_mat_ts$group)) -->
<!-- #  -->
<!-- # tmp <- reshape2::dcast(sel_mat_ts, Gene ~ group, value.var="avg_expression") -->

<!-- # agg_mat <- as.matrix(tmp[, 2:ncol(tmp)]) -->
<!-- # rownames(agg_mat) <- tmp$Gene -->

<!-- agg_mat <- as.matrix(sel_mat) -->

<!-- tmp <- strsplit(colnames(agg_mat), "_") -->
<!-- col_anno <- data.frame(row.names=colnames(agg_mat),  -->
<!--                        Condition = sapply(tmp, `[`,1) |> factor(levels=c("uninfected","Hpy 2m inf")) ,  -->
<!--                        Cluster = sapply(tmp, `[`,2)|> factor(levels=cluster_order)) -->

<!-- agg_mat <- agg_mat[sel_markers_inf, order(col_anno[match(colnames(agg_mat), rownames(col_anno)),"Cluster"],  -->
<!--                            col_anno[match(colnames(agg_mat), rownames(col_anno)),"Condition"])] -->

<!-- anno_cols <- list(Condition=c("uninfected"="lightgreen","Hpy 2m inf"="red")) -->

<!-- pheatmap(agg_mat, cluster_cols=F, cluster_rows = F, scale="row", gaps_col = seq(2, ncol(agg_mat)-2, 2) , breaks=c(-5,-3,seq(-2,2,4/96),3,5),  -->
<!--          annotation_col = col_anno[,c("Condition"), drop=F], -->
<!--          labels_col = col_anno[match(colnames(agg_mat), rownames(col_anno)), "Cluster"], -->
<!--          annotation_colors = anno_cols, -->
<!--          main="Scaled average expression of selected cytokines, Hpy vs. uninfected") -->

<!-- # pheatmap(agg_mat, cluster_cols=F, cluster_rows = F, scale="row", gaps_col = seq(2, ncol(agg_mat)-2, 2) , breaks=c(-5,-3,seq(-2,2,4/96),3,5),  -->
<!-- #          annotation_col = col_anno[,c("Condition"), drop=F], -->
<!-- #          labels_col = col_anno[match(colnames(agg_mat), rownames(col_anno)), "Cluster"], -->
<!-- #          annotation_colors = anno_cols, -->
<!-- #          main="Scaled average expression of selected cytokines, Hpy vs. uninfected",  -->
<!-- #          filename = file.path(result_folder, "Inflammatory cytokines_Hpy_inf_Heatmap.pdf"), width = 8, height = 3) -->

<!-- ``` -->


<!-- ```{r, fig.width=8, fig.height=3} -->
<!-- sel_mat <- AverageExpression(sd_bmp, assays = "RNA", features = sel_markers_inf, group.by = c("condition","cluster_label"))[["RNA"]] -->

<!-- agg_mat <- as.matrix(sel_mat) -->

<!-- tmp <- strsplit(colnames(agg_mat), "_") -->
<!-- col_anno <- data.frame(row.names=colnames(agg_mat),  -->
<!--                        Condition = sapply(tmp, `[`,1) |> factor(levels=c("WT","KO")) ,  -->
<!--                        Cluster = sapply(tmp, `[`,2)|> factor(levels=cluster_order)) -->

<!-- agg_mat <- agg_mat[sel_markers_inf, order(col_anno[match(colnames(agg_mat), rownames(col_anno)),"Cluster"],  -->
<!--                            col_anno[match(colnames(agg_mat), rownames(col_anno)),"Condition"])] -->

<!-- anno_cols <- list(Condition=c("WT"="lightgreen","KO"="red")) -->

<!-- pheatmap(agg_mat, cluster_cols=F, cluster_rows = F, scale="row", gaps_col = seq(2, ncol(agg_mat)-2, 2) , breaks=c(-5,-3,seq(-2,2,4/96),3,5),  -->
<!--          annotation_col = col_anno[,c("Condition"), drop=F], -->
<!--          labels_col = col_anno[match(colnames(agg_mat), rownames(col_anno)), "Cluster"], -->
<!--          annotation_colors = anno_cols, -->
<!--          main="Scaled average expression of selected cytokines, Hpy vs. uninfected") -->

<!-- # pheatmap(agg_mat, cluster_cols=F, cluster_rows = F, scale="row", gaps_col = seq(2, ncol(agg_mat)-2, 2) , breaks=c(-5,-3,seq(-2,2,4/96),3,5),  -->
<!-- #          annotation_col = col_anno[,c("Condition"), drop=F], -->
<!-- #          labels_col = col_anno[match(colnames(agg_mat), rownames(col_anno)), "Cluster"], -->
<!-- #          annotation_colors = anno_cols, -->
<!-- #          main="Scaled average expression of selected cytokines, Bmpr1a KO vs. WT",  -->
<!-- #          filename = file.path(result_folder, "Inflammatory cytokines_Bmpr1a_KO_Heatmap.pdf"), width = 8, height = 3) -->

<!-- ``` -->

<!-- ## Bmp pathway -->

<!-- ```{r, fig.width=10, fig.height=4} -->
<!-- sel_markers_inf <- c("Lgr5","Axin2","Ascl2","Bmpr1a","Id1","Id2","Bmp2","Bmp4","Krt20","Muc5ac","Muc6","Mki67") -->
<!-- p1 <- dotplot_scRNA_multi_feature(sd_inf, cluster = "cluster_label", group = "condition", features = sel_markers_inf, scale=T, group_order = c("uninfected","Hpy 2m inf"), cluster_order = cluster_order) + ggtitle("Regenerative state genes - +/- Hpy infection") + xlab("") + ylab("") -->
<!-- p1 -->

<!-- p2 <- dotplot_scRNA_multi_feature(sd_bmp, cluster = "cluster_label", group = "condition", features = sel_markers_inf, scale=T, group_order = c("WT","KO"), cluster_order = cluster_order) + ggtitle("Inflammatory cytokines - Bmpr1a KO vs. WT") + xlab("") + ylab("") -->
<!-- p2 -->
<!-- ``` -->


<!-- ```{r, fig.width=8, fig.height=6, eval=FALSE} -->
<!-- sel_markers <- c("Bmp7", "Bmp2", "Bmp4", "Bmp5", "Ptger1", "Ptger2", "Ptger3", "Ptger4") -->

<!-- for (gg in sel_markers) { -->
<!--   p <- FeaturePlot(scData.combined, features = gg, order=T, raster = T, pt.size = 3, cols=c("grey80","red"), raster.dpi = c(1000,1000)) -->
<!--   print(p) -->
<!--   pdf(file.path(result_folder, paste0(gg,"_EpiV4.pdf")), width=6, height=5) -->
<!--   print(p) -->
<!--   dev.off() -->
<!-- } -->
<!-- ``` -->


## Stromal markers

### Subpopulation markers from McCarthy et al. (Nat Cell Biol 2020, p.1030 ff)

```{r, fig.width=14, fig.height=4}
sel_markers = unlist(strsplit("Wnt2b, Wnt4, Wnt5a, Rspo1, Rspo2, Rspo3, Dkk2, Dkk3, Sfrp1, Frzb, Wif1, Grem1, Grem2, Chrd, Bmp2, Bmp4, Bmp5, Bmp7, Cd34, Pdgfra, Foxl1, Gli1, Des, Id1, Id2", ", "))
#sel_markers=human2mouse(sel_genes, db=homologeneData2)$mouseGene
DefaultAssay(scData.combined) <- "RNA"

DotPlot(scData.combined, features = sel_markers, scale =T, cols = "RdYlGn", assay = "RNA")  + theme(axis.text.x = element_text(angle=45, hjust=1))

```


<!-- ### Subpopulation markers from Kinchen et al. (Fig. 3) -->

<!-- Str 1: Cxcl12, Adamdec1, Sfrp2, Hcar2, Nxf7,  -->
<!-- Str 2: Wnt5a, Bmp2, Sox6, Fgf9 -->
<!-- Str 3: Maff, Cd55, Pcolce12, Ackr4 -->

<!-- ```{r, fig.width=14, fig.height=6} -->
<!-- sel_markers = c("Cxcl12", "Adamdec1","Pcolce2","Ackr4", "Wnt5a","Bmp2","Sox6", "Fgf9", "Maff","Cd55", "Hcar2", "Nxf7", "Sfrp2") -->
<!-- #sel_markers=human2mouse(sel_genes, db=homologeneData2)$mouseGene -->
<!-- DefaultAssay(scData.combined) <- "RNA" -->

<!-- DotPlot(scData.combined, features = sel_markers, scale =T, cols = "RdYlGn", assay = "RNA") + theme(axis.text.x = element_text(angle=45, hjust=1)) -->
<!-- ``` -->


<!-- ## MF markers -->

<!-- ```{r, fig.width=14, fig.height=6} -->
<!-- sel_markers=human2mouse(c("ACTA2","TAGLN","ACTG2","MYH11"), db=homologeneData2)$mouseGene -->
<!-- DefaultAssay(scData.combined) <- "RNA" -->

<!-- DotPlot(scData.combined, features = sel_markers, scale =T, cols = "RdYlGn", assay = "RNA") + theme(axis.text.x = element_text(angle=45, hjust=1)) -->

<!-- ``` -->


## Proliferation

```{r, fig.width=14, fig.height=4}
sel_markers=c("Top2a","Mki67","Pcna","Mcm3")
DefaultAssay(scData.combined) <- "RNA"

#plot_genes(scData.combined, sel_markers)
DotPlot(scData.combined, features=sel_markers, cols="RdYlBu", assay="RNA")+ theme(axis.text.x=element_text(angle=45, hjust=1))
```




## Mitochondrial transcript proportion

```{r}
FeaturePlot(object = scData.combined,  features = "percent_mito", reduction="umap", label=T)
```


## Scores and RNA count per cluster

```{r, fig.width=14, fig.height=6}
VlnPlot(scData.combined, features = "percent_mito", ncol = 1, pt.size = 0.0)
#VlnPlot(scData.combined, features = "HALLMARK_APOPTOSIS", ncol = 1, pt.size = 0.01)
VlnPlot(scData.combined, features = "nCount_RNA", ncol = 1, pt.size = 0.0) + scale_y_log10()
VlnPlot(scData.combined, features = "nFeature_RNA", ncol = 1, pt.size = 0.0)
```

```{r}
FeaturePlot(object = scData.combined,  features = "nFeature_RNA", reduction="umap", label=T)
```

# Software versions

```{r}
sessionInfo()
```




