---
title: "Cluster identification - Giulia Beccaceci scRNA from stromal Il1r1 KO and epithelial Bmpr1a KO models; epithelial cells only. CCadj, no integration"
author: "Hilmar Berger"
output: 
 html_document:
    toc: true
    toc_depth: 3
    toc_float: false
    number_sections: true
    code_folding: hide
    
pdf_document:
    fig_caption: true
date: "`r format(Sys.time(), '%d %B, %Y %H:%M:%S')`"
---


```{r, message=FALSE}
rm(list=ls())
library(Seurat)
library(ggplot2)

library(SummarizedExperiment)
library(readxl)
library(writexl)
library(knitr)
library(data.table)
library(cowplot)
library(magrittr)
library(dplyr)

library(pheatmap)
library(homologene)

library(grid)
library(png)
library(speckle)
library(knitr)
# library(clusterProfiler)
# 
# options(connectionObserver = NULL)
# library(org.Hs.eg.db)
# library(ReactomePA)
# library(msigdbr)

result_folder = "../../Results/scRNA/noint"
if (!file.exists(result_folder)) dir.create(result_folder, recursive=T)

data_storage_folder = "../../Data/scRNA/Processed"
if (!file.exists(data_storage_folder)) dir.create(data_storage_folder, recursive=T)

library(future)
options("future.globals.maxSize"=2**33)
plan("multisession", workers = 8)

source("../Rutils/Dotplot_scRNA_v2.2.R")

rerun_dge_analyses = T
```

# General

We use preprocessed data that included only subpopulations determined to be epithelial, no doublets. We used variable features only from WT/untreated cells in order to not bias clusters by treatment effects.


```{r}
scData.combined <- readRDS(file.path(data_storage_folder, "Giulia_Il1r1_Bmpr1a_Harmony_Epi_only_v4_CCadj_noint.rds"))
DefaultAssay(scData.combined) <- "RNA"
scData.combined <- JoinLayers(scData.combined)
set.seed(20260710); scData.combined <- FindClusters(scData.combined, resolution = 0.3)
set.seed(1234)
```


```{r}
group_labels <- c("epiBmpr1aKO uninfected" = "Bmpr1a KO","epiBmpr1aWT uninfected"="Bmpr1a WT","Stromal IL1r1 WT Hpy 2m inf"="H.pylori 2m","Stromal IL1r1 WT uninfected"= "Uninfected")

scData.combined@meta.data$group_label <- group_labels[scData.combined$group]
scData.combined@meta.data$Experiment <- ifelse(scData.combined$group_label %in% c("Uninfected","H.pylori 2m"), "Hpy infection", "Bmpr1a KO") |> factor(levels=c("Hpy infection","Bmpr1a KO"))
```


```{r}
group_order = c("Uninfected","H.pylori 2m", "Bmpr1a WT",  "Bmpr1a KO" )
```

```{r}
#scData.combined$cluster_label <- factor(scData.combined$cluster_label, levels = cluster_order)
scData.combined$group_label <- factor(scData.combined$group_label, levels = group_order)
```


```{r, fig.width=7.5, fig.height=5}
meta1 <- readRDS("Results/data_storage/2025-09-10/Giulia_Il1r1_Bmpr1a_Harmony_epi_Metadata_v4.2_CCadj.rds")
scData.combined@meta.data$cluster_label_v4 <- meta1[match(Cells(scData.combined), rownames(meta1)), "cluster_label"]
scData.combined@meta.data$seurat_cluster_v4 <- meta1[match(Cells(scData.combined), rownames(meta1)), "seurat_clusters"]

m <- table(scData.combined$cluster_label_v4, scData.combined$seurat_clusters, dnn=c("integrated","non-integrated"))
pheatmap(m, main = "Comparison of integrated (rows) and non-integrated (columns) clusters")
pheatmap(m, main = "Comparison of integrated (rows) and non-integrated (columns) clusters", filename = file.path(result_folder, "HM_Noint_vs_integrated_all_cells.pdf"))


m <- table(scData.combined$seurat_cluster_v4, scData.combined$seurat_clusters, dnn=c("integrated","non-integrated"))
pheatmap(m, main = "Comparison of integrated (rows) and non-integrated (columns) clusters")


m <- with(subset(scData.combined@meta.data, !cluster_label_v4 %in% c("Squamous","Pit stressed")) |> droplevels(),  table(cluster_label_v4, seurat_clusters, dnn=c("integrated","non-integrated")) )
pheatmap(m, main = "Comparison of integrated (rows) and non-integrated (columns) clusters")
pheatmap(m, main = "Comparison of integrated (rows) and non-integrated (columns) clusters", filename = file.path(result_folder, "HM_non-int_vs_integrated_all_cells.pdf"), width=7.5, height = 5)


```


# Cluster overview

```{r, fig.width=6, fig.height=5}
DimPlot(scData.combined, reduction = "umap", label = TRUE) + NoLegend()
```


```{r, fig.width=12, fig.height=5}
DimPlot(scData.combined, reduction = "umap", label = TRUE, split.by = "group_label", ncol = 4) + NoLegend()
```


# Final cluster assigment

```{r}
cluster_assignment = read.table("Cluster_assignments/Cluster_assignment_Epi_only_2026-07-20_v4_noint.txt", sep="\t",header=T, stringsAsFactors = F)
options(width = 180)
print(cluster_assignment)
rownames(cluster_assignment) = as.character(cluster_assignment$ClusterID)
scData.combined = AddMetaData(scData.combined, cluster_assignment[as.character(scData.combined$seurat_clusters),"Label"], col.name = "cluster_label")
scData.combined = AddMetaData(scData.combined, cluster_assignment[as.character(scData.combined$seurat_clusters),"Type"], col.name = "cluster_type")

scData.combined$cluster_label <- factor(scData.combined$cluster_label, levels = sort(unique(cluster_assignment$Label)))
scData.combined = SetIdent(scData.combined, value="cluster_label" )
```

```{r}
cluster_order <- c("SC antral","Chief","TA antral","TA corpus","Pit/Surface mucous antral","Pit corpus","Epi inflamed","Parietal","G-cell","Tuft","EEC Sst","Enterochromaffine","EC-like","EEC Chga","Squamous")
```


```{r, fig.width=10, fig.height=5}
p = DimPlot(scData.combined, reduction = "umap", label = TRUE) + NoLegend()
print(p)
```


```{r}
scData.combined@meta.data$foo = c(rep("0",100), rep("1",nrow(scData.combined@meta.data)-100))
pdf(file.path(result_folder, "UMAP_annotated_Epi_v4.2_noint.pdf"), width=5, height=5)
p = DimPlot(scData.combined, reduction = "umap", label = TRUE, raster = T, pt.size = 3.5, raster.dpi = c(1000, 1000), shape.by = "foo") + NoLegend()
print(p)
dev.off()
```


```{r}
sd1 <- scData.combined[ , !scData.combined$cluster_label %in% c("Squamous")]
sd1$cluster_label <- droplevels(sd1$cluster_label)
cairo_pdf(file.path(result_folder, "UMAP_annotated_Epi_v4_noint_noraster.pdf"), width=5, height=5)
p = DimPlot(sd1, reduction = "umap", label = TRUE, raster = F, pt.size = 1, raster.dpi = c(1000, 1000)) + NoLegend()
print(p)
dev.off()

cairo_pdf(file.path(result_folder, "UMAP_annotated_Epi_v4_noint_noraster_nolabel.pdf"), width=5, height=5)
p = DimPlot(sd1, reduction = "umap", label = F, raster = F, pt.size = 0.6, raster.dpi = c(1000, 1000)) + NoLegend()
print(p)
dev.off()
```

```{r, fig.width=5, fig.height=5}
cairo_pdf(file.path(result_folder, "UMAP_annotated_Epi_v4.2_noint_noraster_seurat_clusters.pdf"), width=5, height=5)
p = DimPlot(sd1, reduction = "umap", label = TRUE, raster=F, pt.size = .5, group.by="seurat_clusters") + NoLegend()
print(p)
dev.off()
```

```{r, fig.width=15, fig.height=5}
p = DimPlot(scData.combined, reduction = "umap", label = T, split.by="group_label") + NoLegend()
print(p)
```

```{r}
cairo_pdf(file.path(result_folder, "UMAP_annotated_Epi_v4_by_group_noint_noraster_seurat_clusters.pdf"), width=14, height=5)
p = DimPlot(sd1, reduction = "umap", label = F, raster = F, pt.size = 0.6, split.by="group_label", group.by="seurat_clusters") + NoLegend()
print(p)
dev.off()
```

```{r}
cairo_pdf(file.path(result_folder, "UMAP_annotated_Epi_v4_by_group_noint_noraster_nolabel.pdf"), width=14, height=5)
p = DimPlot(sd1, reduction = "umap", label = F, raster = F, pt.size = 0.6, split.by="group_label") + NoLegend()
print(p)
dev.off()
```

## Distribution of clusters within samples

```{r}
m = table(scData.combined$cluster_label, scData.combined$sampleID)
kable(m)
```

```{r, fig.width=10,fig.height=7}
mm  = as.matrix.data.frame(m)
colnames(mm) = colnames(m)
rownames(mm) = rownames(m)

mm_r = sweep(mm, 2, apply(mm,2,sum), "/")

pheatmap(mm_r, cluster_cols = F, cluster_rows=T, main="Proportion of each cluster per sample")
```


## Distribution of clusters within conditions

```{r}
m = table(scData.combined$cluster_label, scData.combined$group_label)
m
```

```{r}
mm  = as.matrix.data.frame(m)
colnames(mm) = colnames(m)
rownames(mm) = rownames(m)

mm_r = sweep(mm, 2, apply(mm,2,sum), "/")

options(width=200)
kable(round(mm_r,3))
```


```{r, fig.width=10,fig.height=7}
pheatmap(mm_r, cluster_cols = F, cluster_rows=T)
```

```{r, fig.width=14, fig.height=8}
mm_rel_ts = reshape2::melt(t(mm_r))
colnames(mm_rel_ts) = c("Group","Cluster","Proportion_cells")
mm_rel_ts$Cluster = factor(paste0("",mm_rel_ts$Cluster))

ggplot(mm_rel_ts, aes(x=Group, y = Proportion_cells, group=Cluster)) + geom_bar(stat="identity") + theme(axis.text.x = element_text(angle=60, hjust=1)) + facet_wrap(~ Cluster )

```

```{r, fig.width=10, fig.height=6}
#cluster_order = sort(unique(mm_rel_ts$Cluster))
#group_order = c("epiBmpr1aWT uninfected","epiBmpr1aKO uninfected","Stromal IL1r1 WT uninfected","Stromal IL1r1 WT Hpy 2m inf")
mm_rel_ts$Cluster = factor(mm_rel_ts$Cluster, levels=cluster_order)
mm_rel_ts$Group <- factor(mm_rel_ts$Group,group_order)
ggplot(mm_rel_ts, aes(x=Group, y = Proportion_cells, fill=Cluster)) + geom_bar(stat="identity") + 
  theme(axis.text.x = element_text(angle=45, hjust=1))
```

```{r, fig.width=12, fig.height=6}
m = table(scData.combined@meta.data$cluster_label, scData.combined@meta.data$orig.ident)
aa <- prop.table(m, margin=2) |> reshape2::melt()
aa <- merge(aa, unique(scData.combined@meta.data[, c("orig.ident","group_label")]), by.x="Var2", by.y="orig.ident")
aa$group <- factor(aa$group_label, levels=group_order)
aa$cluster = factor(aa$Var1, levels=sort(unique(aa$Var1)))
ggplot(aa, aes(x=cluster, fill=group, y=value)) + geom_boxplot() + ggtitle("Clusters in groups") + ylab("Average proportion in samples") + 
  theme(axis.text.x = element_text(angle=45, hjust = 1, size=14))
```

## Comparison of celltype composition

We here use the logit-transformation from R package **speckle** to compare cluster proportions between groups.

```{r, results='asis'}
sd1 <- scData.combined[ , !scData.combined$cluster_label %in% c("Squamous")]
meta <- sd1@meta.data |> droplevels()
m_axin2 <- subset(meta, genotype %in% c("Axin2CreErt2-Bmpr1afl/fl KO","Axin2CreErt2-Bmpr1afl/fl WT"))
res_axin2 <- speckle::propeller(cluster = m_axin2$cluster_label, 
                          sample = m_axin2$orig.ident, 
                          group = as.character(m_axin2$group_label))

m_col1 <- subset(meta, genotype %in% c("Col1a2CreErt2-Il1r1fl/fl WT"))
res_col1 <- speckle::propeller(cluster = m_col1$cluster_label, 
                          sample = m_col1$orig.ident, 
                          group = as.character(m_col1$group_label))

kable(res_axin2)
kable(res_col1)
```

```{r, fig.width=14, fig.height=5}
res_axin2$Experiment <- "Bmpr1a KO"
res_col1$Experiment <- "Hpy infection"
res_comb <- rbind(res_axin2[, c(1:2, 5:9)], res_col1[, c(1:2, 5:9)])
res_comb$cluster <- factor(gsub(" |/" , "\n", res_comb$BaselineProp.clusters), levels=gsub(" |/", "\n", cluster_order))
res_comb$signif <- ifelse(res_comb$P.Value < 0.05, "*", "")
res_comb$pval <- prettyNum(res_comb$P.Value, digits=3)

m = with(sd1@meta.data,table(cluster_label |> droplevels(), orig.ident))
aa <- prop.table(m, margin=2) |> reshape2::melt()
aa <- merge(aa, unique(scData.combined@meta.data[, c("orig.ident","group_label","Experiment")]), by.x="orig.ident", by.y="orig.ident")
aa$group <- factor(aa$group, levels=group_order)

aa$xx <- ifelse(aa$group %in% c("Uninfected","Bmpr1a WT"), 1,2)
aa$cluster = factor(gsub(" |/" , "\n", aa$Var1), levels=gsub(" |/", "\n", cluster_order))
p <- ggplot(aa, aes(x=xx, col=group, y=value)) + geom_point() + stat_summary(geom = "crossbar", fun = mean, width=0.5) + ggtitle("Antral epithelial subcluster proportions across groups - this study") + ylab("Average proportion in samples") + 
  theme(axis.text.x = element_blank(), axis.ticks.x=element_blank(), strip.text = element_text(size=12)) + facet_grid(Experiment ~ cluster, scales = "free_x") + xlab("") + 
  geom_text(data = res_comb, aes(x = 1.5, label = pval), size=4, inherit.aes = F, y = .7) + coord_cartesian(ylim=c(NA, 0.72))

p

cairo_pdf(file.path(result_folder, "Giulia_scRNA_Cell_population_proportions_antral_2rows_v4_noint.pdf"), width=14, height = 5, family="Arial")
print(p)
dev.off()
```

```{r}
saveRDS(scData.combined@meta.data,file=file.path(data_storage_folder, "Giulia_Il1r1_Bmpr1a_Harmony_epi_Metadata_v4_CCadj_noint.rds"))
```

# Marker genes for clusters (assessed on WT cells only)

```{r}
if(rerun_dge_analyses) {
  cluster_markers <- FindAllMarkers(object = scData.combined, only.pos = TRUE, assay = "RNA")
  cluster_markers$diff_pct <- cluster_markers$pct.1 - cluster_markers$pct.2
  saveRDS(cluster_markers, file=file.path(data_storage_folder, "All_cluster_markers_Epi_only_v4_CCadj_noint.rds"))
  write_xlsx(cluster_markers,file.path(result_folder, "All_cluster_markers_Epi_only_v4_CCadj_noint.xlsx"))
} else {
  cluster_markers <- readRDS(file.path(data_storage_folder, "All_cluster_markers_Epi_only_v4_CCadj_noint.rds"))
}
```

```{r}
plot_top_genes <- function(data, dge_results_all, title) {
  for (n in sort(unique(dge_results_all$cluster))) {
    tmp = subset(dge_results_all, cluster==n)
    if(nrow(tmp)<1) next
    tmp_up = subset(tmp, avg_log2FC > 0)
    if(nrow(tmp_up)<1) next
    tmp_up$direction = "up"
    #tmp_down = subset(tmp, avg_log2FC < 0)
    #tmp_down$direction = "down"
    
    #final_top_genes = rbind(head(tmp_up,10), head(tmp_down,10))
    final_top_genes = head(tmp_up, 10)
    final_top_genes = final_top_genes[order(final_top_genes$avg_log2FC),]
    final_top_genes$gene = factor(final_top_genes$gene, levels = final_top_genes$gene)
    p = ggplot(final_top_genes, aes(y=avg_log2FC, x=gene, fill=direction)) + geom_bar(stat="identity") + ggtitle(paste0(n, ", ", title)) + coord_flip() + theme(axis.text.y =  element_text(size=12) )
    print(p)
    
  }
}
```



## Percent expression of top markers across clusters

```{r, fig.width=16, fig.height=14}
top_5_markers = unique(unlist(tapply(cluster_markers$gene, cluster_markers$cluster, head, 5)))
DotPlot(scData.combined, features = top_5_markers , dot.scale = 4, cols = c("red", "blue"), assay = "RNA") +    RotatedAxis()
```



## Log2 Fold Changes of top 10 DE genes from each cluster

```{r, fig.width=7, fig.height=4}
plot_top_genes(scData.combined, cluster_markers, "Cluster markers")
```




```{r}

plot_genes <- function(dataset, genes) {
  
  all_genes =rownames(dataset@assays$RNA)
  
  for (g in genes) {
    if (!g %in% all_genes) {
      print(paste0("Gene ",g," has no data (possibly filtered out)."))
      next
    } else {
        p1  =FeaturePlot(object = dataset, features = g, reduction="umap", label = T)
        p2 = VlnPlot(dataset, features = g, ncol = 1, pt.size = 0.01)
        print(plot_grid(p1,p2, ncol=2,align="h"))
    }
  }
}
```


# Individual marker expression

## Global markers 

These are global cell markers for major cell lineages and types (B/T-cells, Mo/MF etc)

CD45 is Ptprc.  
Pglyrp1 - circulating polymorphonuclear leukocytes (granulocytes + mast cells)  

We here add Cd3e/d/g as pan-T marker and Cd20 (Ms4a1) as mature B-cell marker.
Gata3, Cxcr6, Rora - T-cell markers

Ly6a(Sca-1) - positive in HSC/MPP
Procr,Fdg5, Hoxb5 - HSC

Epor, Slamf1(Cd150) - MegE - lineage

```{r, fig.width=14, fig.height=6}

sel_markers=c("Ptprc","Pglyrp1","Cd4","Cd3e","Cd3d","Cd3g", "Gata3","Cxcr6","Rora","Cd7","Cd8a","Cd19","Ms4a1","Jchain","Aif1","Cd68","Flt3","Ly6a", "Cd34","Procr","Fgd5","Hoxb5","Epor","Slamf1","Vim","Pecam1","Lyve1","Vwf","Col1a1","Myh11","Acta2","Epcam","Krt20","Lgr5","Axin2","S100b","Ncam1","Sox2","Gfap")
DefaultAssay(scData.combined) <- "RNA"

#plot_genes(scData.combined, sel_markers)
DotPlot(scData.combined, features=sel_markers, cols="RdYlBu", assay="RNA")+ theme(axis.text.x=element_text(angle=45, hjust=1))

```

## Granulocyte subpopulations

Pglyrp1 - circulating polymorphonuclear leukocytes (granulocytes + mast cells)  
Csf3r - colony stimulating factor 3 receptor (granulocytes)  G-CSF-R. Higher in mature blood neutrophils (Shinjo et al. 1995 doi: 10.1111/j.1365-2141.1995.tb05390.x)

Mpo,Elane - neutrophil markers  
Prg2 - eosinophils (Dahlin)  
Gzmb, Cma1 - Mast cell (Dahlin)  
Ms4a2, Cpa3 Mast cell, basophil (Dahlin)  
Fcer1a - basophils, mastcells - high affinity IgE receptor  
Ggnt2 - eosinophils  
Il5ra - eosinophil progenitor  



```{r, fig.width=14, fig.height=6}

sel_markers=c("Pglyrp1","Csf3r","Mpo","Elane","Gngt2","Il5ra","Prg2","Gzmb","Cma1","Ms4a2","Cpa3","Fcer1a","Cxcr1","Cxcr2")
DefaultAssay(scData.combined) <- "RNA"

#plot_genes(scData.combined, sel_markers)
DotPlot(scData.combined, features=sel_markers, cols="RdYlBu", assay="RNA")+ theme(axis.text.x=element_text(angle=45, hjust=1))

```


## Monocyte and macrophage markers

Adgre1 = F4/80 = Emr1, Blood Monocytes  

CD14lowCD16high - non-classical MoMF    
CD16 - Fcgr4 (FCGR3A/B)  
Fcgr1 (Cd64) MoMF  
Spn (Cd45) classical Mo  
Cd11b = Itgam  

EMP (Mass et al.) Csf1r + Kit + CD45 low AA4.1(Cd93)+   

```{r, fig.width=14, fig.height=6}
sel_markers=c("Adgre1","Ccr2","Ccr5","Arg1","Mgst1","Fn1","Cd74","Cd68","S100a8","S100a9","Cd14","Fcgr3","Cx3cr1","Fcgr1a","Spn","Itgam","Csf1r","Kit","Cd93","Il7r","Flt3","Cxcr4","Il1b","Aif1","Timd4")
DefaultAssay(scData.combined) <- "RNA"

#plot_genes(scData.combined, sel_markers)
DotPlot(scData.combined, features=sel_markers, cols="RdYlBu", assay="RNA")+ theme(axis.text.x=element_text(angle=45, hjust=1))
```


## DC markers

Xcr1 cDC  
Siglech pDC  
Irf8 pre-DC1  
Ccr7 cDC2  
Pclaf (2810417H13Rik) CMoP / pre-DC1  
Cd209a cDC2, pDC  
Tlr2 cDC  
Tlr4 cDC  

Socs2 Tissue migratory DC (Nirschl et al. 2017, 10.1016/j.cell.2017.06.016)  


```{r, fig.width=14, fig.height=6}
sel_markers=c( "Socs2", "Xcr1", "Siglech", "Cd209a", "Irf8", "Pclaf", "Tlr2","Tlr4","Tlr7","Tlr9","CD1C","CLEC4C")
DefaultAssay(scData.combined) <- "RNA"

#plot_genes(scData.combined, sel_markers)
DotPlot(scData.combined, features=sel_markers, cols="RdYlBu", assay="RNA")+ theme(axis.text.x=element_text(angle=45, hjust=1))
```


## MHC-I/II

```{r, fig.width=14, fig.height=6}
sel_markers=c("H2-Oa","H2-DMa","H2-DMb1","H2-DMb2","H2-Ob","H2-Ab1","H2-Aa","H2-Eb1","H2-Eb2")
#sel_markers = c("HLA-A","HLA-B","HLA-C","HLA-D","HLA-E","HLA-F","HLA-G","HLA-DRA","HLA-DRB","HLA-DQA1","HLA-DQB1","HLA-DPA1","HLA_DPB1","HLA_DMA","HLA_DMB","HLA-DOA","HLA_DOB","MICA","MICB")
DefaultAssay(scData.combined) <- "RNA"

#plot_genes(scData.combined, sel_markers)
DotPlot(scData.combined, features=sel_markers, cols="RdYlBu", assay="RNA")+ theme(axis.text.x=element_text(angle=45, hjust=1))
```

## Stomach site markers

```{r, fig.height=8, fig.width=12}
dotplot_scRNA_multi_feature(scData.combined, features=c("Apc","Axin2","Anxa10","Sox2","Gata4","Meis1","Pdx1","Gata6"),  group="group", cluster="cluster_label", scale=T)+ theme(axis.text.x=element_text(angle=45, hjust=1))
```

```{r}
sd1 <- subset(scData.combined, group %in% c("epiBmpr1aWT uninfected","Stromal IL1r1 WT uninfected"))
```


```{r, fig.height=8, fig.width=10}
dotplot_scRNA_multi_feature(sd1, features=c("Apc","Axin2","Anxa10","Sox2","Gata4","Meis1","Pdx1","Gata6"),  group="tissue_site", cluster="cluster_label", scale=T)+ theme(axis.text.x=element_text(angle=45, hjust=1))
```



```{r, fig.width=16, fig.height=6}
for (gg in c("Apc","Anxa10","Sox2","Gata6","Gata4","Meis1","Pdx1")) {
  p <- FeaturePlot(sd1, features = gg, order=T, split.by="tissue_site") + NoLegend()
  print(p)
}
```

## Stomach epithelial cell markers

```{r, fig.width=14, fig.height=6}
sel_markers=c("Lgr5","Ascl2","Aqp5","Muc5ac","Muc6","Gkn1","Pgc","Pga5", "Cblif","Mki67","Mcm3","Dclk1", "Muc2","Krt20","Olfm4", "Atp4a","Atp4b","Sox2","Cdx2", "Aldh1a1","Tff1","Tff2","Chga","Chgb","Tph1","Sst","Maoa","Maob","Gast","Neurog3","Hdc","Grpr","Dach1","Cntfr","Slc5a5","Adh1c") # |> human2mouse(db=homologeneData2) |> (\(x) x$mouseGene[!is.na(x$mouseGene)] |> unique())()
DefaultAssay(scData.combined) <- "RNA"

#plot_genes(scData.combined, sel_markers)
DotPlot(scData.combined, features=sel_markers, cols="RdYlBu", assay="RNA")+ theme(axis.text.x=element_text(angle=45, hjust=1))
```

## Other epithelial markers

```{r, fig.width=10, fig.height=6}
# ,"Klf4","Myc","Nanog"
DotPlot(scData.combined, c("Sox2","Hnf4a","Hnf4g","Atoh1","Cdx2","Krt5","Krt14","Krt7","Krt19","Krt20","Krt8","Krt18","Aqp8","Car4","Aqp4","Muc5ac","Muc6","Tff1","Tff2","Tff3","Ly6a","Vtn","Gc","Spink4","Muc2"), cols="RdYlBu",assay="RNA") + theme(axis.text.x = element_text(angle=45, hjust=1))
```
```{r, fig.width=10, fig.height=6}
# taken from https://doi.org/10.1016/j.cels.2016.09.002
# Slc18a2 = Vmat2
DotPlot(scData.combined, c("Gcg","Irx2","Arx","Fev","Gc","Ttr","Pgr","Mafb","Ins1","Mafa","Hhex","Sst","Isl1","Rbp4","Gip","Pyy","Etv1","Pax6","Gast","Pdx1","Cck","Nts","Hdc","Tac1","Sct","Neurog3","Olfm1"), cols="RdYlBu",assay="RNA") + theme(axis.text.x = element_text(angle=45, hjust=1))
```



## WAE markers

SSC2c cluster markers from Ayyaz, A., Kumar, S., Sangiorgi, B. et al. Single-cell transcriptomes of the regenerating intestine reveal a revival stem cell. Nature 569, 121–125 (2019). https://doi.org/10.1038/s41586-019-1154-y  

and markers from WAE cells in Nakate et al. (DOI: 10.1126/scitranslmed.adg5252)

```{r, fig.width=8, fig.height=5}
sel_markers = c("Clu","Cd44","Ly6a","Aldh1a3","Cldn4","Krt14","Krt7","Anxa1","Anxa3")
DotPlot(scData.combined, features=sel_markers, cols="RdYlBu", assay="RNA")+ theme(axis.text.x=element_text(angle=45, hjust=1))
```

```{r, fig.width=22, fig.height=8}
sel_markers <- c("Lgr5","Aqp5","Muc5ac","Muc6","Gkn1","Pgc","Cblif","Mki67","Dclk1", "Krt20","Olfm4", "Atp4a","Cdx2","Tff2","Chga","Tph1","Sst","Gast","Cd74","H2-Eb1","Bmpr1a","Il1r1")
dotplot_scRNA_multi_feature(scData.combined, cluster = "cluster_label", group = "group_label", features = sel_markers, scale=T, group_order = group_order)
```

## Inflammation markers / MHC

```{r, fig.width=9, fig.height=8}
sel_markers <- c("Tnf","Cd74","H2-Eb1","H2-Aa","Tap1")
dotplot_scRNA_multi_feature(scData.combined, cluster = "cluster_label", group = "group_label", features = sel_markers, scale=T, group_order = group_order, cluster_order = cluster_order)
```

```{r, fig.width=10, fig.height=4}
FeaturePlot(scData.combined, features = c("Cd74","H2-Eb1","Tnf"), order=T, ncol=3)
```


## Stromal markers

### Subpopulation markers from McCarthy et al. (Nat Cell Biol 2020, p.1030 ff)

```{r, fig.width=14, fig.height=6}
sel_markers = unlist(strsplit("Wnt2b, Wnt4, Wnt5a, Rspo1, Rspo2, Rspo3, Dkk2, Dkk3, Sfrp1, Frzb, Wif1, Grem1, Grem2, Chrd, Bmp2, Bmp4, Bmp5, Bmp7, Cd34, Pdgfra, Foxl1, Gli1, Des, Id1, Id2", ", "))
#sel_markers=human2mouse(sel_genes, db=homologeneData2)$mouseGene
DefaultAssay(scData.combined) <- "RNA"

DotPlot(scData.combined, features = sel_markers, scale =T, cols = "RdYlGn", assay = "RNA")  + theme(axis.text.x = element_text(angle=45, hjust=1))

```


### Subpopulation markers from Kinchen et al. (Fig. 3)

Str 1: Cxcl12, Adamdec1, Sfrp2, Hcar2, Nxf7, 
Str 2: Wnt5a, Bmp2, Sox6, Fgf9
Str 3: Maff, Cd55, Pcolce12, Ackr4

```{r, fig.width=14, fig.height=6}
sel_markers = c("Cxcl12", "Adamdec1","Pcolce2","Ackr4", "Wnt5a","Bmp2","Sox6", "Fgf9", "Maff","Cd55", "Hcar2", "Nxf7", "Sfrp2")
#sel_markers=human2mouse(sel_genes, db=homologeneData2)$mouseGene
DefaultAssay(scData.combined) <- "RNA"

DotPlot(scData.combined, features = sel_markers, scale =T, cols = "RdYlGn", assay = "RNA") + theme(axis.text.x = element_text(angle=45, hjust=1))
```


## MF markers

```{r, fig.width=14, fig.height=6}
sel_markers=human2mouse(c("ACTA2","TAGLN","ACTG2","MYH11"), db=homologeneData2)$mouseGene
DefaultAssay(scData.combined) <- "RNA"

DotPlot(scData.combined, features = sel_markers, scale =T, cols = "RdYlGn", assay = "RNA") + theme(axis.text.x = element_text(angle=45, hjust=1))

```


## Proliferation

```{r, fig.width=14, fig.height=6}
sel_markers=c("Top2a","Mki67","Pcna","Mcm3")
DefaultAssay(scData.combined) <- "RNA"

#plot_genes(scData.combined, sel_markers)
DotPlot(scData.combined, features=sel_markers, cols="RdYlBu", assay="RNA")+ theme(axis.text.x=element_text(angle=45, hjust=1))
```


```{r}
FeaturePlot(scData.combined, "Mki67", order=T)
FeaturePlot(scData.combined, "Mcm3", order=T)
FeaturePlot(scData.combined, "H1f1", order=T)
FeaturePlot(scData.combined, "Top2a", order=T)
```


## Mitochondrial transcript proportion

```{r}
FeaturePlot(object = scData.combined,  features = "percent_mito", reduction="umap", label=T)
```


## Scores and RNA count per cluster

```{r, fig.width=14, fig.height=6}
VlnPlot(scData.combined, features = "percent_mito", ncol = 1, pt.size = 0.0)
#VlnPlot(scData.combined, features = "HALLMARK_APOPTOSIS", ncol = 1, pt.size = 0.01)
VlnPlot(scData.combined, features = "nCount_RNA", ncol = 1, pt.size = 0.0) + scale_y_log10()
VlnPlot(scData.combined, features = "nFeature_RNA", ncol = 1, pt.size = 0.0)
```

```{r}
FeaturePlot(object = scData.combined,  features = "nFeature_RNA", reduction="umap", label=T)
```

<!-- # Marker genes for clusters, healthy only -->

<!-- ```{r} -->
<!-- if(rerun_dge_analyses) { -->
<!--   cluster_markers <- FindAllMarkers(object = subset(scData.combined, condition %in% c("d5_24h_Uninj","d6_48h_Uninj")), only.pos = TRUE, assay = "RNA")  -->
<!--   save(cluster_markers, file=file.path(data_storage_folder, "All_cluster_markers_healthy_all_cells.Rdata")) -->
<!--   write_xlsx(cluster_markers,file.path(result_folder, "All_cluster_markers_healthy_all_cells.xlsx")) -->
<!-- } else { -->
<!--   load(file.path(data_storage_folder, "All_cluster_markers_healthy_all_cells.Rdata")) -->
<!-- } -->
<!-- ``` -->




# Software versions

```{r}
sessionInfo()
```










