---
title: "Final visualizations - Giulia Beccaceci scRNA of epithelial + stromal cells from stromal Il1r1 KO and epithelial Bmpr1a KO models "
author: "Hilmar Berger"
output: 
 html_document:
    toc: true
    toc_depth: 3
    toc_float: false
    number_sections: true
    code_folding: hide
    
pdf_document:
    fig_caption: true
date: "`r format(Sys.time(), '%d %B, %Y %H:%M:%S')`"
---


```{r, message=FALSE}
rm(list=ls())
library(Seurat)
library(ggplot2)

library(SummarizedExperiment)
library(readxl)
library(writexl)
library(knitr)
library(data.table)
library(cowplot)
library(magrittr)
library(dplyr)

library(pheatmap)
library(homologene)

library(grid)
library(png)

# library(clusterProfiler)
# 
# options(connectionObserver = NULL)
# library(org.Hs.eg.db)
# library(ReactomePA)
# library(msigdbr)

source("../Rutils/Dotplot_scRNA_v2.2.R")

result_folder = "../../Results/scRNA/antral_only"
if (!file.exists(result_folder)) dir.create(result_folder, recursive=T)

data_storage_folder = "../../Data/scRNA/Processed"
if (!file.exists(data_storage_folder)) dir.create(data_storage_folder, recursive=T)

library(future)
options("future.globals.maxSize"=2**33)
plan("multisession", workers = 8)
```

# Epithelial cells only

```{r}
scData.combined <- readRDS(file.path(data_storage_folder, "Giulia_Il1r1_Bmpr1a_Harmony_Epi_only_v5_CCadj.rds"))
DefaultAssay(scData.combined) <- "RNA"
scData.combined <- JoinLayers(scData.combined)

set.seed(1234)

meta <- readRDS(file.path(data_storage_folder, "Giulia_Il1r1_Bmpr1a_Harmony_epi_Metadata_v5.1_CCadj.rds"))
scData.combined@meta.data <- meta[match(Cells(scData.combined), rownames(meta)),]

scData.combined <- scData.combined[,(!is.na(scData.combined$cluster_label) & !grepl("stressed|Squamous|Intestinal", scData.combined$cluster_label))]


meta_v4.2 <- readRDS(file.path(data_storage_folder, "Giulia_Il1r1_Bmpr1a_Harmony_epi_Metadata_v4.2_CCadj.rds"))
scData.combined@meta.data$cluster_label_v4.2<- meta_v4.2[match(Cells(scData.combined), rownames(meta_v4.2)), "cluster_label"]
```



```{r}
group_labels <- c("epiBmpr1aKO uninfected" = "Bmpr1a KO","epiBmpr1aWT uninfected"="Bmpr1a WT","Stromal IL1r1 WT Hpy 2m inf"="H.pylori 2m","Stromal IL1r1 WT uninfected"= "Uninfected")

scData.combined@meta.data$group_label <- group_labels[scData.combined$group]
scData.combined@meta.data$Experiment <- ifelse(scData.combined$group_label %in% c("Uninfected","H.pylori 2m"), "Hpy infection", "Bmpr1a KO") |> factor(levels=c("Hpy infection","Bmpr1a KO"))
```


```{r}
group_order = c("Uninfected","H.pylori 2m", "Bmpr1a WT",  "Bmpr1a KO" )
cluster_order <- c("Stem","TA/Progenitor","Pit","Surface mucous","Intestinal-like","stressed cells")
```

```{r}
scData.combined$cluster_label <- factor(scData.combined$cluster_label, levels = cluster_order)
scData.combined$group_label <- factor(scData.combined$group_label, levels = group_order)
```

```{r}
table(scData.combined$orig.ident, scData.combined$group_label)
```


```{r}
table(scData.combined$cluster_label |> droplevels(), scData.combined$group_label)
```

```{r}
de_bmp <- readRDS(file.path(data_storage_folder, "PseudoBulk_DGE_Each_cluster_Axin2CreErt2-Bmpr1a_KO_vs_WT_Epi_only_v5.1_CCadj.rds"))
de_hpy <- readRDS(file.path(data_storage_folder,"PseudoBulk_DGE_Each_cluster_Col1a2CreErt2-Il1r1_Hpy_vs_Uninf_Epi_only_v5.1_CCadj.rds"))

de_res_bmp <- mapply(function(x,n) {x$comparison <- n; x}, de_bmp, names(de_bmp), SIMPLIFY=F) |> do.call(rbind, args = _)
de_res_hpy <- mapply(function(x,n) {x$comparison <- n; x}, de_hpy, names(de_hpy), SIMPLIFY=F) |> do.call(rbind, args = _)
```


## Cluster overview

```{r, fig.width=8, fig.height=5}
p <- DimPlot(scData.combined, reduction = "umap", label = TRUE, group.by="cluster_label") + NoLegend() + ggtitle("UMAP, epithelial cells only")
print(p)
```

```{r}
cairo_pdf(file.path(result_folder, "Giulia_scRNA_UMAP_antral_v5.1.pdf"), width=8, height = 5, family="Arial")
DimPlot(scData.combined, reduction = "umap", label = TRUE, group.by="cluster_label", raster=T, label.size = 5.5) + NoLegend() + ggtitle("UMAP, epithelial cells" )
dev.off()
```

```{r}
cairo_pdf(file.path(result_folder, "S2a_Giulia_scRNA_UMAP_noraster_antral_v5.1.pdf"), width=5, height = 5, family="Arial")
DimPlot(scData.combined, reduction = "umap", label = TRUE, group.by="cluster_label", raster=F, label.size = 4) + NoLegend() + ggtitle("UMAP, epithelial cells")
dev.off()
cairo_pdf(file.path(result_folder, "S2a1_Giulia_scRNA_UMAP_noraster_antral_v5.1_nolabel.pdf"), width=5, height = 5, family="Arial")
DimPlot(scData.combined, reduction = "umap", label = TRUE, group.by="cluster_label", raster=F, label.size = 4) + NoLegend() + ggtitle("UMAP, epithelial cells")
dev.off()
```

```{r, fig.width=14, fig.height=5}
DimPlot(scData.combined, reduction = "umap", label = TRUE, split.by = "group_label", group.by="cluster_label", ncol = 4) + NoLegend() + ggtitle("UMAP all cells, by condition")
```

```{r}
cairo_pdf(file.path(result_folder, "Giulia_scRNA_UMAP_by_group_antral_v5.1.pdf"), width=14, height = 5, family="Arial")
DimPlot(scData.combined, reduction = "umap", label = F, split.by = "group_label", group.by="cluster_label", ncol = 4, raster=T, pt.size = 2, label.size = 5.5) + NoLegend() + ggtitle("UMAP epithelial cells, by condition")
dev.off()
```


```{r}
cairo_pdf(file.path(result_folder, "Giulia_scRNA_UMAP_by_group_noraster_antral_v5.1.pdf"), width=14, height = 5, family="Arial")
DimPlot(scData.combined, reduction = "umap", label = F, split.by = "group_label", group.by="cluster_label", ncol = 4, raster=F, pt.size = 1, label.size = 5.5) + NoLegend() + ggtitle("UMAP epithelial cells, by condition")
dev.off()
```


## Distribution of clusters within conditions


```{r, fig.width=10, fig.height=6}
m = table(scData.combined$cluster_label |> droplevels(), scData.combined$orig.ident)
mm_r <- prop.table(m, margin=2)

mm_rel_ts = reshape2::melt(t(mm_r))
colnames(mm_rel_ts) = c("Sample","Cluster","Proportion_cells")
mm_rel_ts$Cluster = factor(paste0("",mm_rel_ts$Cluster))

tmp <- unique(scData.combined@meta.data[, c("orig.ident", "group_label")])
tmp$Sample_label <- paste0(tmp$group_label, " ", data.table::rowidv(tmp$group_label))
tmp <- tmp[order(tmp$group_label, tmp$Sample_label),]
tmp$Sample_label <- factor(tmp$Sample_label, levels=tmp$Sample_label)
mm_rel_ts <- merge(mm_rel_ts, tmp, by.y="orig.ident", by.x="Sample", all.x=T, sort=F)
mm_rel_ts$Experiment <- ifelse(mm_rel_ts$group_label %in% c("Uninfected","H.pylori 2m"), "Hpy infection", "Bmpr1a KO") |> factor(levels=c("Hpy infection","Bmpr1a KO"))


#cluster_order = sort(unique(mm_rel_ts$Cluster))
mm_rel_ts$Cluster = factor(mm_rel_ts$Cluster, levels=cluster_order) |> droplevels()
#mm_rel_ts$Group <- factor(mm_rel_ts$Group,group_order)
p <- ggplot(mm_rel_ts, aes(x=Sample_label, y = Proportion_cells, fill=Cluster)) + geom_bar(stat="identity") +
  theme(axis.text.x = element_text(angle=45, hjust=1)) + facet_wrap(~Experiment, scale="free_x") + labs(x="", y = "Proportion of cells")
p
```

```{r}
cairo_pdf(file.path(result_folder, "Giulia_scRNA_Cell_population_proportions_barplot_antral_v5.1.pdf"), width=10, height = 6, family="Arial")
print(p)
dev.off()
```



```{r, fig.width=10, fig.height=4}
m = table(scData.combined@meta.data$cluster_label |> droplevels(), scData.combined@meta.data$orig.ident)
aa <- prop.table(m, margin=2) |> reshape2::melt()
aa <- merge(aa, unique(scData.combined@meta.data[, c("orig.ident","group_label")]), by.x="Var2", by.y="orig.ident")
aa$group <- factor(aa$group, levels=group_order)
aa$cluster = factor(gsub(" |/" , "\n", aa$Var1), levels=gsub(" |/", "\n", cluster_order))
p <- ggplot(aa, aes(x=group, col=group, y=value)) + geom_point() + stat_summary(geom = "crossbar", fun = mean, width=0.5) + ggtitle("Epithelial subcluster proportions across groups - this study") + ylab("Average proportion in samples") + 
  theme(axis.text.x = element_blank(), axis.ticks.x=element_blank(), strip.text = element_text(size=12)) + facet_grid(~cluster, scales = "free_x") + xlab("")

p
```

```{r}
cairo_pdf(file.path(result_folder, "Giulia_scRNA_Cell_population_proportions_antral_v5.1.pdf"), width=10, height = 4, family="Arial")
print(p)
dev.off()
```


## Comparison of celltype composition

We here use the logit-transformation from R package **speckle** to compare cluster proportions between groups.

```{r, results='asis'}
meta <- subset(scData.combined@meta.data, !cluster_label %in% c("stressed cells", "Intestinal-like"))
m_axin2 <- subset(meta, genotype %in% c("Axin2CreErt2-Bmpr1afl/fl KO","Axin2CreErt2-Bmpr1afl/fl WT"))
res_axin2 <- speckle::propeller(cluster = droplevels(m_axin2$cluster_label), 
                          sample = m_axin2$orig.ident, 
                          group = as.character(m_axin2$group_label))

m_col1 <- subset(meta, genotype %in% c("Col1a2CreErt2-Il1r1fl/fl WT"))
res_col1 <- speckle::propeller(cluster = droplevels(m_col1$cluster_label), 
                          sample = m_col1$orig.ident, 
                          group = as.character(m_col1$group_label))

kable(res_axin2)
kable(res_col1)
```

```{r, fig.width=7, fig.height=5}
res_axin2$Experiment <- "Bmpr1a KO"
res_col1$Experiment <- "Hpy infection"
res_comb <- rbind(res_axin2[, c(1:2, 5:9)], res_col1[, c(1:2, 5:9)])
res_comb$cluster <- factor(gsub(" |/" , "\n", res_comb$BaselineProp.clusters), levels=gsub(" |/", "\n", cluster_order))
res_comb$signif <- ifelse(res_comb$P.Value < 0.05, "*", "")
res_comb$pval <- prettyNum(res_comb$P.Value, digits=3)

m = with(subset(scData.combined@meta.data, !cluster_label %in% c("stressed cells")),table(cluster_label |> droplevels(), orig.ident))
aa <- prop.table(m, margin=2) |> reshape2::melt()
aa <- merge(aa, unique(scData.combined@meta.data[, c("orig.ident","group_label","Experiment")]), by.x="orig.ident", by.y="orig.ident")
aa$group <- factor(aa$group, levels=group_order)

aa$xx <- ifelse(aa$group %in% c("Uninfected","Bmpr1a WT"), 1,2)
aa$cluster = factor(gsub(" |/" , "\n", aa$Var1), levels=gsub(" |/", "\n", cluster_order))
p <- ggplot(aa, aes(x=xx, col=group, y=value)) + geom_point() + stat_summary(geom = "crossbar", fun = mean, width=0.5) + ggtitle("Antral epithelial subcluster proportions across groups - this study") + ylab("Average proportion in samples") + 
  theme(axis.text.x = element_blank(), axis.ticks.x=element_blank(), strip.text = element_text(size=12)) + facet_grid(Experiment ~ cluster, scales = "free_x") + xlab("") + 
  geom_text(data = res_comb, aes(x = 1.5, label = pval), size=4, inherit.aes = F, y = .7) + coord_cartesian(ylim=c(NA, 0.72))

p

cairo_pdf(file.path(result_folder, "Giulia_scRNA_Cell_population_proportions_antral_2rows_v5.1.pdf"), width=6, height = 5, family="Arial")
print(p)
dev.off()
```

```{r}
tmp <- scData.combined@meta.data
tmp$cluster_label <- factor(tmp$cluster_label, levels=rev(levels(tmp$cluster_label)))
tt <- table(tmp$cluster_label |> droplevels(), tmp$group_label) |> as.data.frame.matrix()
tt
tt <- cbind(Cluster = rownames(tt), tt)
write_xlsx(tt, path=file.path(result_folder, "Epithelial_cell_frequencies_per_group.xlsx"))
```

```{r, fig.width=6.5, fig.height=3.5}

sel_markers=c("Tff1","Tff2","Muc6","Lgr5","Aqp5","Mki67","Mcm3",
              "Muc5ac","Gkn1",
              "Krt20",
              "Cdx2","Olfm4","Muc2")  
DefaultAssay(scData.combined) <- "RNA"

#plot_genes(scData.combined, sel_markers)
p <- DotPlot(scData.combined, features=sel_markers, cols="RdYlBu", assay="RNA", group.by="cluster_label")+ theme(axis.text.x=element_text(angle=45, hjust=1), title = element_text(size=10)) + labs(x="", y="" , title="Selected markers for epithelial cell populations - this study")

p
```

```{r}
cairo_pdf(file.path(result_folder, "Giulia_scRNA_Cluster_markers_antral_v5.1.pdf"), width=6.5, height = 3.5, family="Arial")
print(p)
dev.off()
```


```{r, fig.width=14, fig.height=4}
sel_markers=c("Tff1","Tff2","Muc6","Muc5ac","Pdx1","Sox2","Gkn1","Lgr5","Ascl2","Aqp5","Muc4","Mki67","Mcm3",
              "Gstm3","Sptssb","Aldh3a1","Mogat1",
              "Krt20","Pgc","Pga5", "Cblif","Chia1", "Atp4a","Atp4b",
#              "Krt5","Krt14",
              "Hdc","Chga","Chgb","Neurog3","Sst","Tph1","Gast","Grpr","Dclk1","Krt19","Stmn1","Bmi1","Cckbr","Vegfb","Gper1","Bhlha15","Tnfrsf19")  
DefaultAssay(scData.combined) <- "RNA"

#plot_genes(scData.combined, sel_markers)
p <- DotPlot(scData.combined, features=sel_markers, cols="RdYlBu", assay="RNA", group.by="cluster_label", col.max = 2, col.min = -2)+ theme(axis.text.x=element_text(angle=45, hjust=1)) + labs(x="", y="" , title="Selected markers for epithelial cell populations - this study")

p
```

Science 17 Oct 2024 Vol 386, Issue 6719 pp. 341-348
DOI: 10.1126/science.adl146
The major EEC subtype, the enterochromaffin cell (EC), produces ~90% of the body’s serotonin (5-HT) and regulates gut motility and inflammation. The other EECs are coded with letters: L cells produce glucagon-like peptide 1 (GLP-1), neurotensin (NTS), peptide YY (PYY), and cholecystokinin (CCK); MX cells produce ghrelin (GHRL) and motilin (MLN); D cells produce somatostatin (SST); K cells produce gastric inhibitory protein (GIP); and G cells produce gastrin (GAST) (3). K, MX, and G cells are most abundant in the proximal small intestine (SI), whereas L cells are enriched in the distal SI and colon. The stomach corpus contains an EEC subset called enterochromaffin-like (ECL) cells. In addition to producing some 5-HT, ECL cells produce histamine to regulate acid secretion by nearby parietal cells (4). Important differences exist between mouse and human EECs. For instance, human MX cells express motilin only in the SI but not in the stomach (where these are called X cells), whereas the pertinent gene is a pseudogene in rodents.

```{r, fig.width=14, fig.height=4}
# taken from https://doi.org/10.1016/j.cels.2016.09.002
# Slc18a2 = Vmat2
DotPlot(scData.combined, c("Chga","Chgb","Gcg","Irx2","Arx","Fev","Gc","Ttr","Pgr","Mafb","Ins1","Mafa","Hhex","Sst","Isl1","Rbp4","Gip","Pyy","Etv1","Pax6","Gast","Pdx1","Cck","Nts","Hdc", "Sct","Vtn","Ascl1","Gal","Scgn","Mboat4","Ceacam10","Neurod1","Olfm1","Tac1","Zcchc12","Tead1","Crip1","Peg3"), cols="RdYlBu",assay="RNA", group.by="cluster_label") + theme(axis.text.x = element_text(angle=45, hjust=1))
```

```{r, fig.width=10, fig.height=4}
# taken from https://doi.org/10.1016/j.cels.2016.09.002
# Slc18a2 = Vmat2
DotPlot(scData.combined, c("Pdx1","Sox2","Ascl2","Aqp5","Lgr5","Axin2","Lrig1","Bmi1","Bhlha15","Stmn1","Iqgap3","Tff1","Tff2","Tnfrsf19","Cblif","Chia1","Pga5","Pgc","Atp4a","Muc6","Muc5ac","Krt20","Mki67"), cols="RdYlBu",assay="RNA", group.by="cluster_label", col.max = 1.5, col.min=-1.5) + theme(axis.text.x = element_text(angle=45, hjust=1))
```



## Prostaglandin pathway members

```{r, fig.width=8, fig.height=4}
sel_markers=c("Ptger1","Ptger2","Ptger3","Ptger4")
DefaultAssay(scData.combined) <- "RNA"

#plot_genes(scData.combined, sel_markers)
p <- DotPlot(scData.combined, features=sel_markers, cols="RdYlBu", assay="RNA", group = "cluster_label", dot.scale = 8)+ theme(axis.text.x=element_text(angle=45, hjust=1)) + xlab("") + ylab("") + ggtitle("Prostaglandin pathway, epithelial cells - this study")
p
```

```{r}
cairo_pdf(file.path(result_folder, "Giulia_scRNA_Prostaglandins_antral_v5.1.pdf"), width=8, height = 4, family="Arial")
print(p)
dev.off()
```

```{r, fig.width=12, fig.height=6}
sel_markers=c("Ptger1","Ptger2","Ptger3","Ptger4")
DefaultAssay(scData.combined) <- "RNA"

#plot_genes(scData.combined, sel_markers)
p <- dotplot_scRNA_multi_feature(scData.combined, features=sel_markers, assay="RNA", group = "group_label", cluster="cluster_label", 
                            group_order = group_order, cluster_order = cluster_order,
                            scale=T) + xlab("") + ylab("") + ggtitle("Prostaglandin pathway, epithelial cells - this study")
p
```

```{r}
cairo_pdf(file.path(result_folder, "Giulia_scRNA_Prostaglandins_by_group_antral_v5.1.pdf"), width=12, height = 4, family="Arial")
print(p)
dev.off()
```

```{r, fig.width=12, fig.height=5}
VlnPlot(scData.combined, features = "Ptger4", group.by="cluster_label", split.by="group_label") + xlab("")
```


```{r, fig.width=12, fig.height=9}
de_res_sel <- subset(de_res_bmp, GeneSymbol == "Ptger4")
rownames(de_res_sel) <- NULL
comp_suffix <- "_KO_vs_WT"
de_res_sel$cluster <- gsub(paste0(comp_suffix,"$"),"", de_res_sel$comparison)
de_res_sel$pval <- prettyNum(de_res_sel$pvalue, digits=3)
de_res_sel$padj_t <- prettyNum(ifelse(is.na(de_res_sel$padj), 1, de_res_sel$padj), digits=3)
#de_res_sel$ypos <- max(de_res_sel$baseMean, na.rm=T) * 1.1

p1 <- VlnPlot(subset(scData.combined, Experiment == "Bmpr1a KO"), features = "Ptger4", group.by="cluster_label", split.by="group_label", y.max = 3.1) + xlab("") + 
  geom_text(data = de_res_sel, aes(x = cluster, y = 3, label = padj_t), inherit.aes = F, size = 3) 

de_res_sel <- subset(de_res_hpy, GeneSymbol == "Ptger4")
rownames(de_res_sel) <- NULL
comp_suffix <- "_Hp_vs_ui"
de_res_sel$cluster <- gsub(paste0(comp_suffix,"$"),"", de_res_sel$comparison)
de_res_sel$pval <- prettyNum(de_res_sel$pvalue, digits=3)
de_res_sel$padj_t <- prettyNum(ifelse(is.na(de_res_sel$padj), 1, de_res_sel$padj), digits=3)
#de_res_sel$ypos <- max(de_res_sel$baseMean, na.rm=T) * 1.1

p2 <- VlnPlot(subset(scData.combined, Experiment == "Hpy infection"), features = "Ptger4", group.by="cluster_label", split.by="group_label", y.max = 3.1) + xlab("") +
  geom_text(data = de_res_sel, aes(x = cluster, y = 3, label = padj_t), inherit.aes = F, size = 3) 

plot_grid(p1, p2, nrow=2, ncol=1)

cairo_pdf(file.path(result_folder, "Ptger4_Violin_antral_v5.1.pdf"), width=12, height = 9, family="Arial")
plot_grid(p1, p2, nrow=2, ncol=1)
dev.off()
```


## Il1 and Il1 receptors

```{r, fig.width=6, fig.height=3.5}
sel_markers=c("Il1a","Il1b","Il1r1","Il1r2","Il1rap")
DefaultAssay(scData.combined) <- "RNA"

#plot_genes(scData.combined, sel_markers)
p <- DotPlot(scData.combined, features=sel_markers, cols="RdYlBu", assay="RNA", group = "cluster_label", dot.scale = 8)+ theme(axis.text.x=element_text(angle=45, hjust=1), title = element_text(size=11)) + xlab("") + ylab("") + ggtitle("Il1, epithelial cells - this study")
p
```


```{r}
cairo_pdf(file.path(result_folder, "Giulia_scRNA_Il1_antral_v5.1.pdf"), width=6, height = 3.5, family="Arial")
print(p)
dev.off()
```

```{r, fig.width=9, fig.height=3.5}
sel_markers=c("Il1a","Il1b","Il1r1","Il1r2","Il1rap")
DefaultAssay(scData.combined) <- "RNA"

#plot_genes(scData.combined, sel_markers)
p <- dotplot_scRNA_multi_feature(scData.combined, features=sel_markers, assay="RNA", group = "group_label", cluster="cluster_label", 
                            group_order = group_order, cluster_order = cluster_order,
                            scale=T) + xlab("") + ylab("") + ggtitle("Il1, epithelial cells - this study")
p
```

```{r}
cairo_pdf(file.path(result_folder, "Giulia_scRNA_Il1_by_group_antral_v5.1.pdf"), width=9, height = 3.5, family="Arial")
print(p)
dev.off()
```


## Cytokine Violin plots

```{r, fig.width=8, fig.height=6}
VlnPlot(scData.combined, features = c("Cxcl1","Ccl20"), ncol=1, group.by="cluster_label", split.by="group_label")
```

```{r, fig.width=14, fig.height=3.5}
#sel_features <- c("Cxcl1","Cxcl2","Cxcl3","Cxcl5","Ccl20","Cxcl10","Tnfsf4","Alpk1","Isg15","Socs1")
sel_features <- c("Cxcl1","Cxcl2","Cxcl3","Cxcl5","Ccl20","Cxcl10","Alpk1","Tnfsf4","Isg15","Socs1")
p <- dotplot_scRNA_multi_feature(scData.combined, features = sel_features, cluster="cluster_label", group="group_label", scale=T, max_val_scale = 2, min_val_scale = -2) + labs(x="", y="", title="Cytokine expression in epithelial cells, this study")
p

cairo_pdf(file.path(result_folder, "Giulia_scRNA_Cytokines_and_ALPK1_by_group_antral_v5.1.pdf"), width=14, height = 3.5, family="Arial")
print(p)
dev.off()
```

# Signatures on Epithelial cells - UCell

```{r, fig.width = 20, fig.height=12}
library(UCell)
library(GSEABase)
library(homologene)
library(msigdbr)

ee = msigdbr(category = "H", species = "mouse")

mouse_gs <- GSEABase:::getGmt("../../Data/External/GeneSets/mouse_custom_geneset_2026-04-24.gmt")
human_gs <- GSEABase:::getGmt("../../Data/External/GeneSets/human_custom_geneset_2026-04-24.gmt")

#gene_sets <- list()
gene_sets <- tapply(ee$gene_symbol, ee$gs_name, list)

sel_mouse_gs <- names(mouse_gs)
sel_mouse_gs <- sel_mouse_gs [!grepl("Mouse_colon_ref|AKP_", sel_mouse_gs)]

sel_human_gs <- names(human_gs)
sel_human_gs <- sel_human_gs [!grepl("Joanito|Cords|Gao_FB|Qi_CRC|Tissue_specific|Healthy_human_stomach", sel_human_gs)]

for (nn in sel_mouse_gs) {
  gene_sets[[make.names(nn)]] <- geneIds(mouse_gs[[nn]])   
}

for (nn in sel_human_gs) {
    gene_sets[[make.names(nn)]] <- geneIds(human_gs[[nn]]) |> homologene::human2mouse(db=homologene::homologeneData2) |> (\(x) x$mouseGene)()
}

set.seed(1234)
scData.combined <- UCell::AddModuleScore_UCell(scData.combined, features = gene_sets)
```

```{r}
tmp <- scData.combined@meta.data[, grepl("_UCell", colnames(scData.combined@meta.data))] |> as.matrix()
rownames(tmp) <- rownames(scData.combined@meta.data)
colnames(tmp) <- gsub("_UCell","", colnames(tmp))
r <- t(tmp)

```


```{r}
library(limma)
inp_mat <- r
sid <- scData.combined@meta.data[colnames(inp_mat),"orig.ident"]
cl <- scData.combined@meta.data[colnames(inp_mat),"cluster_label"]
group <- scData.combined@meta.data[colnames(inp_mat),"group_label"]

ed <- data.frame(sample=as.character(sid), g1 =group |> as.character() |> make.names()) |> unique()
rownames(ed) <- ed$sample

all_res <- list()

for (cc in unique(cl)) {
  r_agg <- limma::avearrays(inp_mat[,cl == cc], sid[cl==cc])

  library(limma)
  #design <- cbind(intercept=rep(1,ncol(r_agg)), group=ifelse(ed[colnames(r_agg),"g1"]=="AKP", 1, 0) )

  mm <- model.matrix(~g1+0, data = ed[match(colnames(r_agg), ed$sample),])
  
  fit <- lmFit(r_agg, design = mm)
  
  if(length(colnames(mm)) == 4) {
  cont.matrix <- makeContrasts(Bmpr1a_KO_vs_WT = g1Bmpr1a.KO - g1Bmpr1a.WT,
                              Hpy_vs_uninf =    g1H.pylori.2m - g1Uninfected,
                              levels=mm)
  } else {
    if( all(c("g1Bmpr1aKO","g1Bmpr1aWT") %in% colnames(mm)) ) {
        cont.matrix <- makeContrasts(Bmpr1a_KO_vs_WT = g1Bmpr1a.KO - g1epiBmpr1a.WT,
                              levels=mm)
    } else { 
      if( all(c("g1H.pylori.2m","g1Uninfected") %in% colnames(mm)) ) {
        cont.matrix <- makeContrasts(Hpy_vs_uninf =    g1H.pylori.2m - g1Uninfected,
                              levels=mm)
      } else {
        next;
      }
    }
  }
  
  
  fit2 <- contrasts.fit(fit, cont.matrix)
  fit2 <- eBayes(fit2)
  # Bmpr1a_KO_vs_WT <- topTable(fit2, coef = "Bmpr1a_KO_vs_WT", adjust="BH", number = nrow(r_agg))
  # Hpy_vs_uninf <- topTable(fit2, coef = "Hpy_vs_uninf", adjust="BH", number = nrow(r_agg))
  
  for (nn in colnames(cont.matrix)) {
    res <- topTable(fit2, coef = nn, adjust="BH", number = nrow(r_agg))
    res$gs_name <- rownames(res)
    res$cluster <- cc
    res$comparison <- nn
    res$clust_comp <- paste0(cc, "__", nn)
    all_res[[paste0(cc, "__", nn)]] <- res[order(res$P.Value),]
  }

}


all_res_df <- do.call(rbind, all_res)
all_res_df$FDR.global  <- p.adjust(all_res_df$P.Value, method = "BH")
```


```{r, fig.width=12, fig.height=5}
sel_features <- c("Gregorieff_Yap_KO_down_UCell", "Gregorieff_Yap_Tg_up_UCell", "Yui_Fetal_vs_adult_up_UCell","Gil_Vazquez_RSC_signature_mouse_UCell" ,
                 "Vallone_Trop2_Signature_UCell","Wang_Yap_signature_UCell", "HALLMARK_INFLAMMATORY_RESPONSE_UCell", "HALLMARK_TNFA_SIGNALING_VIA_NFKB_UCell", "HALLMARK_OXIDATIVE_PHOSPHORYLATION_UCell","HALLMARK_FATTY_ACID_METABOLISM_UCell")
DotPlot(scData.combined, features = sel_features, group.by="cluster_label", scale=T, cols="RdYlBu") + theme(axis.text.x=element_text(angle=45, hjust=1))
```


```{r}
sel_clusters <- unique(scData.combined$cluster_label) |> as.character()
sel_clusters <- sel_clusters[!grepl("EEC|EC|Tuft|G-cell", sel_clusters)]
tmp <- scData.combined@meta.data |> subset(cluster_label %in% sel_clusters) |> as.data.table()
```

```{r}
de <- all_res_df
```


```{r, fig.width=18, fig.height=7}
sel_genesets <- c("Gil_Vazquez_RSC_signature_mouse_UCell", "Yui_Fetal_vs_adult_up_UCell","Yui_DSS_colitis_vs_control_up_UCell", "HALLMARK_INFLAMMATORY_RESPONSE_UCell","Gregorieff_Yap_KO_down_UCell","Gregorieff_Yap_Tg_up_UCell", "Lgr5_SC_Signature_Munoz_UCell", "HALLMARK_OXIDATIVE_PHOSPHORYLATION_UCell","HALLMARK_FATTY_ACID_METABOLISM_UCell")

for (gs in sel_genesets) {
  gn <- gsub("_UCell","", gs)
  
  p <- ggplot(tmp, aes(x=group_label, fill=group_label, y = .data[[gs]])) + geom_boxplot() + 
  ggpubr::stat_compare_means(comparisons = list(c( "Uninfected","H.pylori 2m"), c( "Bmpr1a WT", "Bmpr1a KO" )), method = "wilcox") + facet_wrap(~ cluster_label, ncol=11) +
  theme(axis.text.x=element_text(angle=45, hjust=1))
  print(p)
  
  tmp_agg <- tmp[, .(mean_score = mean(.SD[[gs]]) ), by=c("orig.ident", "group_label", "cluster_label")]
  tmp_agg$Experiment <- ifelse(tmp_agg$group_label %in% c("Uninfected","H.pylori 2m"), "Hpy infection", "Bmpr1a KO") |> factor(levels=c("Hpy infection","Bmpr1a KO"))
  
  de_gs <- subset(de, gs_name == gn)
  de_gs$Experiment <- ifelse(de_gs$comparison == "Bmpr1a_KO_vs_WT", "Bmpr1a KO", "Hpy infection")
  de_gs$sign <- ifelse(de_gs$adj.P.Val < 0.05, "*", "n.s.")
  de_gs$cluster_label <- factor(de_gs$cluster, levels=cluster_order)
  de_gs <- subset(de_gs, cluster_label %in% tmp_agg$cluster_label)
  de_gs$adjp <- prettyNum(de_gs$adj.P.Val, digits=3)
  de_gs$pval <- prettyNum(de_gs$P.Value, digits=3)
  de_gs$ypos <- max(tmp_agg$mean_score, na.rm=T) * 1.1
  
  p <- ggplot(tmp_agg, aes(x=group_label, col=group_label, y = mean_score)) + 
      #geom_boxplot() + 
      stat_summary(geom = "crossbar", fun = mean, width=0.5) + geom_point() +
      geom_text(data = de_gs, aes(x = 1.5, y = ypos, label = pval), inherit.aes = F) +
      facet_wrap(Experiment ~ cluster_label, ncol=11, scale = "free_x") + 
      theme(axis.text.x=element_text(angle=45, hjust=1)) + 
      ggtitle(paste0("Average ",gn," UCell score per sample")) + labs(x="",  y= "Avg UCell score") 
  print(p )
  
  cairo_pdf(file.path(result_folder, paste0(gn, "_signature_UCell_antral_v5.1.pdf")), width=14, height = 7, family="Arial")
  print(p)
dev.off()


}
```


```{r, fig.width=9, fig.height=12}
sel_genesets <- c("Gil_Vazquez_RSC_signature_mouse_UCell", 
                  "HALLMARK_INFLAMMATORY_RESPONSE_UCell", 
                  "Gregorieff_Yap_KO_down_UCell", 
                  "Yui_Fetal_vs_adult_up_UCell" )

for (gg in sel_genesets) {
  p1 <- FeaturePlot(subset(scData.combined, Experiment=="Hpy infection"), features = gg, order=T, split.by="group_label", max.cutoff = "q90", min.cutoff = "q50", cols=c("grey","red")) + scale_y_continuous(sec.axis = sec_axis(transform= ~., name=""))
  p2 <- FeaturePlot(subset(scData.combined, Experiment=="Bmpr1a KO"), features = gg, order=T, split.by="group_label", max.cutoff = "q90", min.cutoff = "q50", cols=c("grey","red")) + scale_y_continuous(sec.axis = sec_axis(transform= ~., name=""))
  print(plot_grid(p1,p2, nrow=2, ncol=1, labels=gg, labels.size=.5)  )
  
  cairo_pdf(file.path(result_folder, paste0(gg, "_FeaturePlot_antral_v5.1.pdf")), width=9, height = 9, family="Arial")
  print(plot_grid(p1,p2, nrow=2, ncol=1))
  dev.off()
}
```


```{r, fig.width=12, fig.height=9}
p1 <- VlnPlot(subset(scData.combined, Experiment == "Bmpr1a KO"), features = "Gil_Vazquez_RSC_signature_mouse_UCell", group.by="cluster_label", split.by="group_label") + xlab("")
p2 <- VlnPlot(subset(scData.combined, Experiment == "Hpy infection"), features = "Gil_Vazquez_RSC_signature_mouse_UCell", group.by="cluster_label", split.by="group_label") + xlab("")

plot_grid(p1, p2, nrow=2, ncol=1)

cairo_pdf(file.path(result_folder, "Gil_Vazquez_RSC_signature_mouse_UCell_Violin_antral_v5.1.pdf"), width=12, height = 9, family="Arial")
plot_grid(p1, p2, nrow=2, ncol=1)
dev.off()
```

```{r, fig.width=12, fig.height=9}
p1 <- VlnPlot(subset(scData.combined, Experiment == "Bmpr1a KO"), features = "HALLMARK_INFLAMMATORY_RESPONSE_UCell", group.by="cluster_label", split.by="group_label") + xlab("")
p2 <- VlnPlot(subset(scData.combined, Experiment == "Hpy infection"), features = "HALLMARK_INFLAMMATORY_RESPONSE_UCell", group.by="cluster_label", split.by="group_label") + xlab("")

plot_grid(p1, p2, nrow=2, ncol=1)

cairo_pdf(file.path(result_folder, "HALLMARK_INFLAMMATORY_RESPONSE_UCell_Violin_antral_v5.1.pdf"), width=12, height = 9, family="Arial")
plot_grid(p1, p2, nrow=2, ncol=1)
dev.off()
```


# Final figures 3b and S1e/f (Heatmaps)


```{r}
#cluster_order2 <- c("SC antral","Chief-Pro","Antral-pro Cldn2 Muc4","Chief/Parietal","TA antral","TA corpus","Pit antral early","Pit antral","Pit corpus","Surface mucus Krt20","Parietal","G-cell","Tuft","EEC Sst","EC-like cell","EEC Chga","EEC Tph1","EEC Neurog3")

sd_bmp <- subset(scData.combined, genotype %in% c("Axin2CreErt2-Bmpr1afl/fl KO","Axin2CreErt2-Bmpr1afl/fl WT"))
sd_bmp@meta.data$condition <- gsub("epiBmpr1a","", sd_bmp$group) |> gsub(" uninfected", "", x=_)

sd_inf <- subset(scData.combined, genotype %in% c("Col1a2CreErt2-Il1r1fl/fl WT"))
sd_inf@meta.data$condition <- gsub("Stromal IL1r1 WT ","", sd_inf$group)
```

## Fetal/Yap targets

```{r, fig.width=14, fig.height=3.5}
sel_markers <- c("Tead4", "Wfdc2", "Itga2", "Tgm2", "Lrg1", "Anxa1", "Sftpd", "Ccn1", "Vnn1", "Clu", "Socs3", "Ly6a", "Il33", "Ly6d", "Gpnmb", "Ereg")

p1 <- dotplot_scRNA_multi_feature(sd_inf, cluster = "cluster_label", group = "condition", features = sel_markers, scale=T, group_order = c("uninfected","Hpy 2m inf"), cluster_order = cluster_order) + ggtitle("Regenerative state genes - +/- Hpy infection") + xlab("") + ylab("")
p1
p2 <- dotplot_scRNA_multi_feature(sd_bmp, cluster = "cluster_label", group = "condition", features = sel_markers, scale=T, group_order = c("WT","KO"), cluster_order = cluster_order) + ggtitle("Regenerative state genes - KO vs. WT") + xlab("") + ylab("")
p2

pdf(file.path(result_folder, "Selected_Regenerative_state_genes_Hpy_inf_antral_v5.1.pdf"), width=14, height = 3.5)
print(p1)
dev.off()

pdf(file.path(result_folder, "Selected_Regenerative_state_genes_Bmpr1a_KO_antral_v5.1.pdf"), width=14, height = 3.5)
print(p2)
dev.off()

```

```{r}
make_star_mat <- function(template_map, de_res, all_clusters, g1, g2, comp_suffix, p_thr = 0.05) {
  #stars <- matrix("", nrow = nrow(agg_mat), ncol = ncol(agg_mat),
#                dimnames = dimnames(agg_mat))
#stars[pval < 0.05] <- "*"
de_res_sel <- subset(de_res, GeneSymbol %in% rownames(template_map))
rownames(de_res_sel) <- NULL
de_res_sel$cluster <- gsub(paste0(comp_suffix,"$"),"", de_res_sel$comparison)
de_res_sel <- subset(de_res_sel, cluster %in% all_clusters)
missing_cl <- as.character(all_clusters[!all_clusters %in% unique(de_res_sel$cluster)])
if(!length(missing_cl)==0) {
  tmp_row <- de_res_sel[1, ,drop=F]
  tmp_row$padj = 1
  tmp_row$pvalue = 1
  tmp_row$log2FoldChange = NA
  for (cl in missing_cl) {
    tmp_row$cluster <- cl
    de_res_sel <- rbind(de_res_sel, tmp_row)
  }
}
de_res_sel$signif <- ifelse(de_res_sel$padj  <= p_thr, "*", "")
tmp <- reshape2::dcast(de_res_sel, GeneSymbol ~ cluster, value.var="signif")
de_res_star_mat <- as.matrix(tmp[, 2:ncol(tmp)])
rownames(de_res_star_mat) <- tmp$GeneSymbol
tmp2 <- de_res_star_mat
tmp2[tmp2!=""] <- ""
colnames(de_res_star_mat) <- paste0(g1,"_", colnames(de_res_star_mat))
colnames(tmp2) <- paste0(g2,"_", colnames(tmp2))
de_res_star_mat <- cbind(de_res_star_mat, tmp2)

de_res_star_mat <- de_res_star_mat[rownames(template_map), colnames(template_map)]
de_res_star_mat[is.na(de_res_star_mat)] <- ""
return(de_res_star_mat)
}
```



```{r, fig.width=8.5, fig.height=5}
sel_mat <- AverageExpression(sd_inf, assays = "RNA", features = sel_markers, group.by = c("condition","cluster_label"))[["RNA"]]


agg_mat <- as.matrix(sel_mat)

tmp <- strsplit(colnames(agg_mat), "_")
col_anno <- data.frame(row.names=colnames(agg_mat), 
                       Condition = sapply(tmp, `[`,1) |> factor(levels=c("uninfected","Hpy 2m inf")) , 
                       Cluster = sapply(tmp, `[`,2)|> factor(levels=cluster_order))

agg_mat <- agg_mat[sel_markers, order(col_anno[match(colnames(agg_mat), rownames(col_anno)),"Cluster"], 
                           col_anno[match(colnames(agg_mat), rownames(col_anno)),"Condition"])]

all_clusters <- unique(sd_inf$cluster_label) |> as.character()
star_mat <- make_star_mat(agg_mat, de_res_hpy, all_clusters, g1 = "Hpy 2m inf", g2="uninfected",comp_suffix = "_Hp_vs_ui", p_thr = 0.1 )

anno_cols <- list(Condition=c("uninfected"="lightgreen","Hpy 2m inf"="red"))

pheatmap(agg_mat, cluster_cols=F, cluster_rows = F, scale="row", gaps_col = seq(2, ncol(agg_mat)-2, 2) , breaks=c(-5,-3,seq(-1,1,2/96),3,5), 
         annotation_col = col_anno[,c("Condition"), drop=F],
         labels_col = col_anno[match(colnames(agg_mat), rownames(col_anno)), "Cluster"],
         annotation_colors = anno_cols,
         main="Scaled average expression of regenerative markers, Hpy vs. uninfected", 
         display_numbers = star_mat,
         number_color = "black",
         fontsize_number = 14)

pheatmap(agg_mat, cluster_cols=F, cluster_rows = F, scale="row", gaps_col = seq(2, ncol(agg_mat)-2, 2) , breaks=c(-5,-3,seq(-1,1,2/96),3,5), 
         annotation_col = col_anno[,c("Condition"), drop=F],
         labels_col = col_anno[match(colnames(agg_mat), rownames(col_anno)), "Cluster"],
         annotation_colors = anno_cols,
         main="Scaled average expression of regenerative markers, Hpy vs. uninfected", 
         display_numbers = star_mat,
         number_color = "black",
         fontsize_number = 14,
         filename = file.path(result_folder, "S1e - Regenerative_state_markers_Hpy_inf_Heatmap_antral_v5.1.pdf"), width = 5, height = 5)

pheatmap(agg_mat, cluster_cols=F, cluster_rows = F, scale="row", gaps_col = seq(2, ncol(agg_mat)-2, 2) , breaks=c(-5,-3,seq(-1,1,2/96),3,5), 
         annotation_col = col_anno[,c("Condition"), drop=F],
         labels_col = col_anno[match(colnames(agg_mat), rownames(col_anno)), "Cluster"],
         annotation_colors = anno_cols,
         main="Scaled average expression of regenerative markers, Hpy vs. uninfected", 
         fontsize_number = 14,
         filename = file.path(result_folder, "S1e1 - Regenerative_state_markers_Hpy_inf_Heatmap_antral_v5.1_nostars.pdf"), width = 5, height = 5)

```


```{r, fig.width=8.5, fig.height=5}
sel_mat <- AverageExpression(sd_bmp, assays = "RNA", features = sel_markers, group.by = c("condition","cluster_label"))[["RNA"]]

agg_mat <- as.matrix(sel_mat)

tmp <- strsplit(colnames(agg_mat), "_")
col_anno <- data.frame(row.names=colnames(agg_mat), 
                       Condition = sapply(tmp, `[`,1) |> factor(levels=c("WT","KO")) , 
                       Cluster = sapply(tmp, `[`,2)|> factor(levels=cluster_order))

agg_mat <- agg_mat[sel_markers, order(col_anno[match(colnames(agg_mat), rownames(col_anno)),"Cluster"], 
                           col_anno[match(colnames(agg_mat), rownames(col_anno)),"Condition"])]

all_clusters <- unique(sd_inf$cluster_label) |> as.character()
star_mat <- make_star_mat(agg_mat, de_res_bmp, all_clusters, g1 = "KO", g2="WT",comp_suffix = "_KO_vs_WT", p_thr = 0.1 )


anno_cols <- list(Condition=c("WT"="lightgreen","KO"="red"))

pheatmap(agg_mat, cluster_cols=F, cluster_rows = F, scale="row", gaps_col = seq(2, ncol(agg_mat)-2, 2) , breaks=c(-5,-3,seq(-1,1,2/96),3,5), 
         annotation_col = col_anno[,c("Condition"), drop=F],
         labels_col = col_anno[match(colnames(agg_mat), rownames(col_anno)), "Cluster"],
         annotation_colors = anno_cols,
         main="Scaled average expression of regenerative markers, Bmpr1a KO vs. WT",
         display_numbers = star_mat,
         number_color = "black",
         fontsize_number = 14)

pheatmap(agg_mat, cluster_cols=F, cluster_rows = F, scale="row", gaps_col = seq(2, ncol(agg_mat)-2, 2) , breaks=c(-5,-3,seq(-1,1,2/96),3,5), 
         annotation_col = col_anno[,c("Condition"), drop=F],
         labels_col = col_anno[match(colnames(agg_mat), rownames(col_anno)), "Cluster"],
         annotation_colors = anno_cols,
         main="Scaled average expression of regenerative markers, Bmpr1a KO vs. WT", 
         display_numbers = star_mat,
         number_color = "black",
         fontsize_number = 14,
         filename = file.path(result_folder, "S1f - Regenerative_state_markers_Bmpr1a_KO_Heatmap_antral_v5.1.pdf"), width = 5, height = 5)

pheatmap(agg_mat, cluster_cols=F, cluster_rows = F, scale="row", gaps_col = seq(2, ncol(agg_mat)-2, 2) , breaks=c(-5,-3,seq(-1,1,2/96),3,5), 
         annotation_col = col_anno[,c("Condition"), drop=F],
         labels_col = col_anno[match(colnames(agg_mat), rownames(col_anno)), "Cluster"],
         annotation_colors = anno_cols,
         main="Scaled average expression of regenerative markers, Bmpr1a KO vs. WT", 
         fontsize_number = 14,
         filename = file.path(result_folder, "S1f - Regenerative_state_markers_Bmpr1a_KO_Heatmap_antral_v5.1_nostars.pdf"), width = 5, height = 5)
```

## Cytokines

```{r, fig.width=8, fig.height=3.5, eval=FALSE}
sel_markers_inf <- c("Cxcl1","Cxcl2","Cxcl5","Ccl20","Cxcl10")
p1 <- dotplot_scRNA_multi_feature(sd_inf, cluster = "cluster_label", group = "condition", features = sel_markers_inf, scale=T, group_order = c("uninfected","Hpy 2m inf"), cluster_order = cluster_order) + ggtitle("Inflammatory cytokines - +/- Hpy infection") + xlab("") + ylab("")
p1

p2 <- dotplot_scRNA_multi_feature(sd_bmp, cluster = "cluster_label", group = "condition", features = sel_markers_inf, scale=T, group_order = c("WT","KO"), cluster_order = cluster_order) + ggtitle("Inflammatory cytokines - Bmpr1a KO vs. WT") + xlab("") + ylab("")
p2

pdf(file.path(result_folder, "Inflammatory cytokines_Hpy_inf_antral_v5.1.pdf"), width=8, height = 3.5)
print(p1)
dev.off()

pdf(file.path(result_folder, "Inflammatory cytokines_Bmpr1a_KO_antral_v5.1.pdf"), width=8, height = 3.5)
print(p2)
dev.off()
```

```{r, fig.width=8, fig.height=3}
sel_markers_inf <- c("Cxcl1","Cxcl2","Cxcl5","Ccl20","Cxcl10")
sel_mat <- AverageExpression(sd_inf, assays = "RNA", features = sel_markers_inf, group.by = c("condition","cluster_label"))[["RNA"]]

agg_mat <- as.matrix(sel_mat)

tmp <- strsplit(colnames(agg_mat), "_")
col_anno <- data.frame(row.names=colnames(agg_mat), 
                       Condition = sapply(tmp, `[`,1) |> factor(levels=c("uninfected","Hpy 2m inf")) , 
                       Cluster = sapply(tmp, `[`,2)|> factor(levels=cluster_order))

agg_mat <- agg_mat[sel_markers_inf, order(col_anno[match(colnames(agg_mat), rownames(col_anno)),"Cluster"], 
                           col_anno[match(colnames(agg_mat), rownames(col_anno)),"Condition"])]

all_clusters <- unique(sd_inf$cluster_label) |> as.character()
star_mat <- make_star_mat(agg_mat, de_res_hpy, all_clusters, g1 = "Hpy 2m inf", g2="uninfected",comp_suffix = "_Hp_vs_ui", p_thr = 0.1 )

anno_cols <- list(Condition=c("uninfected"="lightgreen","Hpy 2m inf"="red"))

pheatmap(agg_mat, cluster_cols=F, cluster_rows = F, scale="row", gaps_col = seq(2, ncol(agg_mat)-2, 2) , breaks=c(-5,-3,seq(-2,2,4/96),3,5), 
         annotation_col = col_anno[,c("Condition"), drop=F],
         labels_col = col_anno[match(colnames(agg_mat), rownames(col_anno)), "Cluster"],
         annotation_colors = anno_cols,
         main="Scaled average expression of selected cytokines, Hpy vs. uninfected",
         display_numbers = star_mat,
         number_color = "black",
         fontsize_number = 14)

pheatmap(agg_mat, cluster_cols=F, cluster_rows = F, scale="row", gaps_col = seq(2, ncol(agg_mat)-2, 2) , breaks=c(-5,-3,seq(-2,2,4/96),3,5), 
         annotation_col = col_anno[,c("Condition"), drop=F],
         labels_col = col_anno[match(colnames(agg_mat), rownames(col_anno)), "Cluster"],
         annotation_colors = anno_cols,
         main="Scaled average expression of selected cytokines, Hpy vs. uninfected", 
         display_numbers = star_mat,
         number_color = "black",
         fontsize_number = 14,
         filename = file.path(result_folder, "3b - Inflammatory cytokines_Hpy_inf_Heatmap_antral_v5.1.pdf"), width = 5, height = 3)

pheatmap(agg_mat, cluster_cols=F, cluster_rows = F, scale="row", gaps_col = seq(2, ncol(agg_mat)-2, 2) , breaks=c(-5,-3,seq(-2,2,4/96),3,5), 
         annotation_col = col_anno[,c("Condition"), drop=F],
         labels_col = col_anno[match(colnames(agg_mat), rownames(col_anno)), "Cluster"],
         annotation_colors = anno_cols,
         main="Scaled average expression of selected cytokines, Hpy vs. uninfected", 
         display_numbers = star_mat,
         number_color = "black",
         fontsize_number = 14,
         filename = file.path(result_folder, "3b1 - Inflammatory cytokines_Hpy_inf_Heatmap_antral_v5.1_nostars.pdf"), width = 5, height = 3)


```


```{r, fig.width=8, fig.height=3}
sel_mat <- AverageExpression(sd_bmp, assays = "RNA", features = sel_markers_inf, group.by = c("condition","cluster_label"))[["RNA"]]

agg_mat <- as.matrix(sel_mat)

tmp <- strsplit(colnames(agg_mat), "_")
col_anno <- data.frame(row.names=colnames(agg_mat), 
                       Condition = sapply(tmp, `[`,1) |> factor(levels=c("WT","KO")) , 
                       Cluster = sapply(tmp, `[`,2)|> factor(levels=cluster_order))

agg_mat <- agg_mat[sel_markers_inf, order(col_anno[match(colnames(agg_mat), rownames(col_anno)),"Cluster"], 
                           col_anno[match(colnames(agg_mat), rownames(col_anno)),"Condition"])]

all_clusters <- unique(sd_inf$cluster_label) |> as.character()
star_mat <- make_star_mat(agg_mat, de_res_bmp, all_clusters, g1 = "KO", g2="WT",comp_suffix = "_KO_vs_WT", p_thr = 0.1 )


anno_cols <- list(Condition=c("WT"="lightgreen","KO"="red"))

pheatmap(agg_mat, cluster_cols=F, cluster_rows = F, scale="row", gaps_col = seq(2, ncol(agg_mat)-2, 2) , breaks=c(-5,-3,seq(-2,2,4/96),3,5), 
         annotation_col = col_anno[,c("Condition"), drop=F],
         labels_col = col_anno[match(colnames(agg_mat), rownames(col_anno)), "Cluster"],
         annotation_colors = anno_cols,
         main="Scaled average expression of selected cytokines, Hpy vs. uninfected",
         display_numbers = star_mat,
         number_color = "black",
         fontsize_number = 14)

pheatmap(agg_mat, cluster_cols=F, cluster_rows = F, scale="row", gaps_col = seq(2, ncol(agg_mat)-2, 2) , breaks=c(-5,-3,seq(-2,2,4/96),3,5), 
         annotation_col = col_anno[,c("Condition"), drop=F],
         labels_col = col_anno[match(colnames(agg_mat), rownames(col_anno)), "Cluster"],
         annotation_colors = anno_cols,
         main="Scaled average expression of selected cytokines, Bmpr1a KO vs. WT", 
         display_numbers = star_mat,
         number_color = "black",
         fontsize_number = 14,
         filename = file.path(result_folder, "3b - Inflammatory cytokines_Bmpr1a_KO_Heatmap_antral_v5.1.pdf"), width = 5, height = 3)

pheatmap(agg_mat, cluster_cols=F, cluster_rows = F, scale="row", gaps_col = seq(2, ncol(agg_mat)-2, 2) , breaks=c(-5,-3,seq(-2,2,4/96),3,5), 
         annotation_col = col_anno[,c("Condition"), drop=F],
         labels_col = col_anno[match(colnames(agg_mat), rownames(col_anno)), "Cluster"],
         annotation_colors = anno_cols,
         main="Scaled average expression of selected cytokines, Bmpr1a KO vs. WT", 
         fontsize_number = 14,
         filename = file.path(result_folder, "3b - Inflammatory cytokines_Bmpr1a_KO_Heatmap_antral_v5.1_nostars.pdf"), width = 5, height = 3)
```

## Bmp pathway

```{r, fig.width=10, fig.height=6, eval=FALSE}
sel_markers_inf <- c("Lgr5","Axin2","Ascl2","Bmpr1a","Id1","Id2","Bmp2","Bmp4","Krt20","Muc5ac","Muc6","Mki67")
p1 <- dotplot_scRNA_multi_feature(sd_inf, cluster = "cluster_label", group = "condition", features = sel_markers_inf, scale=T, group_order = c("uninfected","Hpy 2m inf"), cluster_order = cluster_order) + ggtitle("Regenerative state genes - +/- Hpy infection") + xlab("") + ylab("")
p1

p2 <- dotplot_scRNA_multi_feature(sd_bmp, cluster = "cluster_label", group = "condition", features = sel_markers_inf, scale=T, group_order = c("WT","KO"), cluster_order = cluster_order) + ggtitle("Inflammatory cytokines - Bmpr1a KO vs. WT") + xlab("") + ylab("")
p2
```


# UMAPs for Bmps2/4/5/7 and Ptger1-4


```{r, fig.width=6, fig.height=3.5}
sel_markers <- c("Bmp2", "Bmp7", "Bmp4", "Bmp5")

p <- DotPlot(scData.combined, features = sel_markers, scale=T, cols="RdYlBu", assay="RNA", group.by="cluster_label") + theme(axis.text.x=element_text(angle=45, hjust=1)) + xlab("") + ylab("")
p
cairo_pdf(file.path(result_folder, "S2ix - Bmp Dotplot global_antral_v5.1.pdf"), width=6, height=3.5, family="Arial")
print(p)
dev.off()

```

```{r, fig.width=8, fig.height=3.5}
p <- dotplot_scRNA_multi_feature(scData.combined, features = sel_markers, scale=T, assay="RNA", cluster="cluster_label", group="group_label")
p
cairo_pdf(file.path(result_folder, "S2ix - Bmp Dotplot per_group_antral_v5.1.pdf"), width=8, height=3.5, family="Arial")
print(p)
dev.off()
```


```{r, fig.width=18, fig.height=3.5}
all_plots <- list()
for (gg in sel_markers) {
  p <- FeaturePlot(scData.combined, features = gg, order=T, raster = T, pt.size = 3, cols=c("grey80","red"), raster.dpi = c(1000,1000), max.cutoff = "q90")
  all_plots[[gg]] <- p
}

cairo_pdf(file.path(result_folder, "S2i - Bmp UMAPs_antral_v5.1.pdf"), width=18, height=3.5, family="Arial")
print(plot_grid(plotlist = all_plots, ncol=4, nrow=1))
dev.off()
```

```{r, fig.width=8, fig.height=6}
sel_markers <- c("Ptger1", "Ptger2", "Ptger3", "Ptger4")

all_plots <- list()
for (gg in sel_markers) {
  p <- FeaturePlot(scData.combined, features = gg, order=T, raster = T, pt.size = 3, cols=c("grey80","red"), raster.dpi = c(1000,1000), max.cutoff = "q90")
  all_plots[[gg]] <- p
}

cairo_pdf(file.path(result_folder, "S6e - Ptger UMAPs_antral_v5.1.pdf"), width=18, height=4, family="Arial")
print(plot_grid(plotlist = all_plots, ncol=4, nrow=1))
dev.off()
```

# Selected regeneration target genes  - Violin plots 

```{r, eval=FALSE}
sel_markers <- c("Lgr5","Anxa1", "Id1", "Ccn1", "Ankrd1", "Ccn2")
all_plots <- list()
for (g in sel_markers) {
  p <- VlnPlot(scData.combined, features = g, group.by="cluster_label", split.by="group_label")
  print(p)
  all_plots[[g]] <- p
}

cairo_pdf(file.path(result_folder, "Rev3_sel_markers_Violin_antral_v5.1.pdf"), height=18, width=6, family="Arial")
print(plot_grid(plotlist = all_plots, ncol=1))
dev.off()
```


```{r, fig.width=12, fig.height=9}
vln_with_p <- function(ds, gene, de_res, suffix, y_text = 3) {
  de_res_sel <- subset(de_res, GeneSymbol == gene)
  rownames(de_res_sel) <- NULL
  comp_suffix <- suffix
  de_res_sel$cluster <- gsub(paste0(comp_suffix,"$"),"", de_res_sel$comparison)
  de_res_sel$pval <- prettyNum(de_res_sel$pvalue, digits=3)
  de_res_sel$padj_t <- prettyNum(ifelse(is.na(de_res_sel$padj), 1, de_res_sel$padj), digits=3)
  #de_res_sel$ypos <- max(de_res_sel$baseMean, na.rm=T) * 1.1
  de_res_sel <- subset(de_res_sel, cluster %in% unique(ds$cluster_label))
  p1 <- VlnPlot(ds, features = gene, group.by="cluster_label", split.by="group_label", y.max = 3.1) + 
    geom_text(data = de_res_sel, aes(x = cluster, label = padj_t),y = y_text,  inherit.aes = F, size = 3) + 
    xlab("") + 
    coord_cartesian(ylim=c(NA, y_text))
  return(p1)
}
```

```{r, fig.width=6, fig.height=8}
sel_markers <- c("Lgr5","Anxa1", "Id1","Id2", "Ccn1", "Ccn2")
all_plots <- list()
for (g in sel_markers) {
  #p1 <- VlnPlot(sd_inf, features = g, group.by="cluster_label", split.by="group_label")
  #p2 <- VlnPlot(sd_bmp, features = g, group.by="cluster_label", split.by="group_label")
  p1 <- vln_with_p(sd_inf, g, de_res_hpy, "_Hp_vs_ui", y_text = 3.2)
  p2 <- vln_with_p(sd_bmp, g, de_res_bmp, "_KO_vs_WT", y_text = 3.2)
  px <- plot_grid(p1,p2,ncol=2, nrow=1)
  print(px)
  all_plots[[g]] <- px
}

cairo_pdf(file.path(result_folder, "Rev3_sel_markers_Violin_antral_v5.1.pdf"), height=24, width=12, family="Arial")
print(plot_grid(plotlist = all_plots, ncol=1))
dev.off()
```

# Software versions

```{r}
sessionInfo()
```

