---
title: "Giulia Beccaceci scRNA from stromal Il1r1 KO and epithelial Bmpr1a KO models - Preprocessing of epithelial cells only"
author: "Hilmar Berger"
output: 
 html_document:
    toc: true
    toc_depth: 3
    toc_float: false
    number_sections: true
    code_folding: hide
    
pdf_document:
    fig_caption: true
date: "`r format(Sys.time(), '%d %B, %Y %H:%M:%S')`"
---

```{r, message=FALSE}
library(Seurat)
#library(SeuratDisk)
library(ggplot2)

library(readxl)
library(knitr)
library(data.table)
library(cowplot)
library(magrittr)

library(harmony)

library(homologene)

library(pheatmap)

library(plotly)

library(future)
plan("multicore", workers = 10)
options(future.globals.maxSize = 10*2**32)

data_storage_folder = file.path("../../Data/scRNA/Processed")
if (!file.exists(data_storage_folder)) dir.create(data_storage_folder, recursive=T)

set.seed(20230829)

```

We keep only cells from epithelial cells that were not marked as doublets. 

```{r}
reload_data = F
if(reload_data) {
  load(file.path(data_storage_folder, "All_Filtered_scData.Rdata"))
  
  meta <- readRDS(file=file.path(data_storage_folder, "Giulia_Il1r1_Bmpr1a_Harmony_Metadata.rds"))
  
  all_ge_data_filtered@meta.data <- meta[Cells(all_ge_data_filtered),]
  
  good_cells <- subset(meta, cluster_type=="Epithelial" & doublet=="singlet" & !grepl("decaying", cluster_label, ignore.case = T))
  all_ge_data_filtered <- subset(all_ge_data_filtered, cells = rownames(good_cells))
  all_ge_data_filtered$cluster_label <- droplevels(all_ge_data_filtered$cluster_label)
}
```

# Cluster counts from global clustering

```{r}
if(reload_data) {
  table(all_ge_data_filtered$cluster_label, all_ge_data_filtered$sampleID)
}  
```


# Cell Cycle estimation

```{r}
cc_estimate_file = file.path(data_storage_folder, "Giulia_epi_only_CellCycle_estimated.rds")
if(reload_data) {

  scData = all_ge_data_filtered
  scData <- NormalizeData(object = scData, normalization.method = "LogNormalize", scale.factor = 1e4)
  scData <- FindVariableFeatures(object = scData, selection.method = "vst")
  scData <- ScaleData(object = scData, features = rownames(scData))

  scData <- RunPCA(scData, features = VariableFeatures(scData), ndims.print = c(1:20), nfeatures.print = 20)

  mouse.cc.genes = lapply(cc.genes.updated.2019, function(x) unique(human2mouse(x, db = homologeneData2)$mouseGene))
  #scData <- CellCycleScoring(scData, s.features = s.genes, g2m.features = g2m.genes, set.ident = TRUE)
  scData <- CellCycleScoring(scData, s.features = mouse.cc.genes$s.genes, g2m.features = mouse.cc.genes$g2m.genes, set.ident = TRUE)

  cc_tab = scData@meta.data[, c("Phase","S.Score","G2M.Score", "orig.ident")]
  #rm(scData)
  #gc()
  saveRDS(cc_tab, file=cc_estimate_file, compress = T)
} else {
  cc_tab = readRDS(cc_estimate_file)
}
```



```{r}
# view cell cycle scores and phase assignments
boxplot(cc_tab$S.Score ~ cc_tab$Phase)
boxplot(cc_tab$G2M.Score ~ cc_tab$Phase)

m = table(cc_tab$Phase, cc_tab$orig.ident)
kable(m)

tmp = as.matrix.data.frame(m)
colnames(tmp) = colnames(m)
rownames(tmp) = rownames(m)
mr = sweep(tmp, 2, apply(tmp,2,sum),"/")
kable(round(mr,2))

pheatmap(mr, scale="none", main="Relative contributions of cell cycle phases per sample", breaks=seq(0,1,by=0.01))
```



We here use standard normalization followed by batch integration using Harmony (without adjusting for cell cycle or MT content). 

```{r}
if(reload_data) {
  combined_sc_data <- NormalizeData(object = all_ge_data_filtered, normalization.method = "LogNormalize", scale.factor = 1e4)
  combined_sc_data <- FindVariableFeatures(object = combined_sc_data, selection.method = "vst")
}
```

```{r}
if(reload_data) {
  # restrict variable features to those variable within WT untreated samples  
  tmp <- subset(combined_sc_data, group %in% c("Stromal IL1r1 WT uninfected","epiBmpr1aWT uninfected"))
  tmp <- FindVariableFeatures(object = tmp, selection.method = "vst")
  tmp_var <- VariableFeatures(tmp)
  combined_sc_data@meta.data$S.Score <- cc_tab[match(Cells(combined_sc_data), rownames(cc_tab)), "S.Score"]
  combined_sc_data@meta.data$G2M.Score <- cc_tab[match(Cells(combined_sc_data), rownames(cc_tab)), "G2M.Score"]  
  combined_sc_data <- ScaleData(object = combined_sc_data, features = tmp_var, vars.to.regress = c("S.Score","G2M.Score"))
  combined_sc_data <- RunPCA(combined_sc_data, verbose = FALSE, features = tmp_var)
  combined_sc_data <- RunHarmony(object = combined_sc_data,
                                    reduction.use = "pca",
                                    dims.use = 1:50,
                                    group.by.vars = "orig.ident",
                                    plot_convergence = TRUE)
  combined_sc_data <- RunUMAP(combined_sc_data, reduction = "harmony", dims = 1:15, verbose = FALSE)
  combined_sc_data <- RunUMAP(combined_sc_data, reduction = "harmony", dims = 1:15, verbose = FALSE, reduction.name = "umap3", n.components = 3 )
  combined_sc_data <- FindNeighbors(combined_sc_data, reduction = "harmony", dims = 1:15)
  combined_sc_data <- FindClusters(combined_sc_data, resolution = 0.65)
  scData.combined = combined_sc_data
}
```


```{r} 
preprocessed_data_file = file.path(data_storage_folder, "Giulia_Il1r1_Bmpr1a_Harmony_Epi_only_v4_CCadj.rds")
if(reload_data) {
  saveRDS(scData.combined, file=preprocessed_data_file, compress = T)
} else {
  scData.combined <- readRDS(preprocessed_data_file)
}
```


# Global clusters

```{r, fig.width=14, fig.height=6}
scData.combined <- SetIdent(scData.combined, value="seurat_clusters")
DimPlot(scData.combined, reduction = "umap", group.by = "orig.ident", raster = T)

p1 <- DimPlot(scData.combined, reduction = "umap", group.by = c("genotype", "tissue_type"), raster=T)
p2 <- DimPlot(scData.combined, reduction = "umap", label = TRUE, raster=T)
plot_grid(p1, p2)

DimPlot(scData.combined, reduction = "umap", split.by = "orig.ident", label = T, raster=T)
DimPlot(scData.combined, reduction = "umap", group.by="cluster_label", label = T, raster=T)
```


```{r}
d <- as.data.frame(scData.combined@reductions$umap3@cell.embeddings)
d <- merge(d, scData.combined@meta.data, by=0, all.x=T, sort=F)
fig <-  plot_ly(data = d ,x =  ~umap3_1, y = ~umap3_2, z = ~umap3_3, marker = list(size = 3) ) %>% 
  add_markers(color = ~cluster_label) %>%
  layout( 
    xaxis = list(
      zerolinecolor = "#ffff",
      zerolinewidth = 2,
      gridcolor='#ffff'), 
    yaxis = list(
      zerolinecolor = "#ffff",
      zerolinewidth = 2,
      gridcolor='#ffff'),
    scene =list(bgcolor = "#e5ecf6"),
    legend=list(itemsizing = "constant"))
fig

fig <-  plot_ly(data = d ,x =  ~umap3_1, y = ~umap3_2, z = ~umap3_3,  marker = list(size = 1)) %>% 
  add_markers(color = ~group) %>%
  layout( 
    xaxis = list(
      zerolinecolor = "#ffff",
      zerolinewidth = 2,
      gridcolor='#ffff'), 
    yaxis = list(
      zerolinecolor = "#ffff",
      zerolinewidth = 2,
      gridcolor='#ffff'),
    scene =list(bgcolor = "#e5ecf6"),
    legend=list(itemsizing = "constant"))
fig

```

## Cluster assignments

```{r}
m = table(scData.combined@meta.data$seurat_clusters, scData.combined@meta.data$orig.ident)
kable(m)
```


```{r, fig.height=6, fig.width=12}
mm = as.matrix.data.frame(m)
colnames(mm) = colnames(m)
rownames(mm) = rownames(m)
mm_rel = sweep(mm, 2, apply(mm, 2,sum), "/")
pheatmap(mm_rel, main="Relative single cell contribution of each cluster to condition totals")
```


# Global cluster labels

```{r}
table(scData.combined$cluster_label, scData.combined$seurat_clusters)
```


## Global markers

```{r, fig.width=14, fig.height=6}

sel_markers=c("Ptprc","Tcrb","Ly6g","Pglyrp1","Cd4","Cd3e","Cd3d","Cd3g", "Gata3","Cxcr6","Rora","Cd7","Cd8a","Cd19","Ms4a1","Ly6a", "Cd34","Procr","Fgd5","Hoxb5","Epor","Slamf1", "Vim","Epcam","Pecam1","Lyve1","Vwf","Mki67")
DefaultAssay(scData.combined) <- "RNA"

DotPlot(scData.combined, sel_markers, cols = "RdYlBu", assay = "RNA", scale=T) + theme(axis.text.x=element_text(angle=45, hjust = 1))

```



```{r, fig.width=10, fig.height=5}
VlnPlot(scData.combined, features = c("nFeature_RNA", "percent_mito"))
```


# Software versions

```{r}
sessionInfo()
```

