---
title: "Microarray Data - Bmpr1a KO - Final Volcano plot"
author: "Hilmar Berger"
output: 
 html_document:
    toc: true
    toc_depth: 3
    toc_float: false
    number_sections: true
    code_folding: hide
    
pdf_document:
    fig_caption: true
date: "`r format(Sys.time(), '%d %B, %Y %H:%M:%S')`"
---

```{r, messages=FALSE}
rm(list=ls())
library(limma)
library(readxl)
library(pheatmap)
library(xtable)
library(reshape2)
library(knitr)
library(writexl)
library(EnhancedVolcano)
library(data.table)

#load("../../Data/Processed/Micro_array_preprocessed_data.Rdata")

result_folder = "../../Results"
if (!file.exists(result_folder)) dir.create(result_folder, recursive=T)
```

```{r}

filename = paste(result_folder,"DGE_results.Rdata",sep="/")
load(filename)
```


## Volcano plots

Due to the small replicate number no genes in any of the comparisons reached a significance level of FDR < 5%. We here therefore show unadjusted (raw) p-values noting that **a large part or all of them might be false positive calls**.

The raw p-value threshold is 0.001. We only keep the probe with the significant one with highest absolute value or (if none is significant) the one with highest absolute log2FC overall. We also exclude probes mapping to multiple genes. 

```{r, volcano, fig.width=8, fig.height=6}
all_target_conditions = names(all_results)
par(mfrow=c(1,1), mar=c(8,4,4,2))

sel_genes = "Cxcl2,Cxcl10,S100a8,Cxcl6,Ccl20,Il1b,S100a9,Il1a,Il6,Ccl2,C4b,C3" |> strsplit(",") |> unlist()


get_max_fc <- function(x) {
  tmp <- x
  tmp$psig <- tmp$P.Value < 0.001
  setorder(tmp, -psig, -abs_lfc)
  tmp[1, ]$abs_lfc
}

for (tc in all_target_conditions) {
  r = all_results[[tc]] |> as.data.table()
  r <- r[r$GencodeM22MappingStatus %in% c("unique gene,perfect match", "no perfect matching probe") & !is.na(r$GeneSymbol) & r$GeneSymbol != ""]
  r$abs_lfc <- abs(r$logFC)
  r[, max_sign_lfc := get_max_fc(.SD), by="GeneSymbol"]
  
  r2 <- r[r$max_sign_lfc == r$abs_lfc]
  
  p = EnhancedVolcano(r2, lab=r2$GeneSymbol, x="logFC", y="P.Value", pCutoff = 0.001, 
                        selectLab = sel_genes, max.overlaps=Inf, drawConnectors = T, subtitle = "")
  print(p)
}
```


# Session Info
```{r}
sessionInfo()
```